---
title: "RNA-Seq_coexpression_analysis"
author: "Neha Mishra"
date: "2026-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Learning objectives

This tutorial's central theme is **sex-sensitive analysis**: most coexpression tutorials treat sex as, at best, a nuisance covariate to adjust away. Here it's the opposite — sex is a variable we investigate on purpose, at every stage, because a coexpression program (or a module's relationship to disease severity) can genuinely look different in males and females, and averaging over that difference can hide real biology or mislead about what's driving a signal.

By the end of this tutorial you will be able to:

-   Explain what a gene coexpression network is and why WGCNA is a useful tool to build one
-   Go from a normalized expression matrix to a set of coexpression **modules**
-   Choose a soft-thresholding power using the scale-free topology criterion
-   Summarize each module with a **module eigengene** and interpret it
-   Test whether modules are associated with clinical/phenotypic traits, with mixed models that account for repeated timepoints per patient
-   **Test whether a module-trait relationship differs by sex** (a statistical interaction), and distinguish that from simply adjusting for sex as a covariate
-   (Optional, advanced) Build separate coexpression networks per sex, and formally test whether modules discovered in one sex are structurally preserved in the other

> This notebook expects an `input/` folder (sample metadata, gene annotation, clinical data) and an `output_QC/` folder (normalized counts) in the same directory as this `.Rmd` file — knitr uses the file's own folder as the working directory, so no `setwd()` is needed. A new `output_coexpression/` folder is created automatically to hold this tutorial's outputs.

> 🧪 boxes mark a short exercise — try them before moving to the next chunk.

## 0. Why analyze coexpression networks by sex?

Sex differences in immune and inflammatory response are well documented (e.g. in COVID-19 severity and outcomes), yet many transcriptomic studies either exclude sex from the model entirely or only ever include it as an additive covariate — which corrects for a shift in the *average* level of a module, but can't detect or describe a difference in how strongly (or in which direction) a module tracks disease.

Throughout this tutorial you'll build one shared network first (the standard workflow), then progressively ask sharper sex-specific questions of it: does sex shift a module's baseline level (a covariate), does the *relationship* between a module and a clinical trait differ by sex (an interaction), and finally, does the *network structure itself* — which genes cluster together at all — differ by sex (a structural, not just statistical, question). 

## 1. Install packages

Run this once per machine. Each line only installs a package if it isn't already present, so it's safe to re-run.

```{r install-packages, message=FALSE, results='hide'}
# general
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("tidyverse", quietly = TRUE))
    install.packages("tidyverse")
if (!requireNamespace("reshape2", quietly = TRUE))
    install.packages("reshape2")
if (!requireNamespace("matrixStats", quietly = TRUE))
    install.packages("matrixStats")

# network analysis
if (!requireNamespace("WGCNA", quietly = TRUE))
    BiocManager::install("WGCNA")

# statistics
if (!requireNamespace("lme4", quietly = TRUE))
    install.packages("lme4")
if (!requireNamespace("lmerTest", quietly = TRUE))
    install.packages("lmerTest")

# functional enrichment
if (!requireNamespace("topGO", quietly = TRUE))
    BiocManager::install("topGO")
if (!requireNamespace("org.Hs.eg.db", quietly = TRUE))
    BiocManager::install("org.Hs.eg.db")

# visualizations
if (!requireNamespace("ComplexHeatmap", quietly = TRUE))
    BiocManager::install("ComplexHeatmap")
if (!requireNamespace("RColorBrewer", quietly = TRUE))
    install.packages("RColorBrewer")
if (!requireNamespace("circlize", quietly = TRUE))
    install.packages("circlize")
if (!requireNamespace("corrplot", quietly = TRUE))
    install.packages("corrplot")
if (!requireNamespace("viridis", quietly = TRUE))
    install.packages("viridis")
if (!requireNamespace("cowplot", quietly = TRUE))
    install.packages("cowplot")
if (!requireNamespace("patchwork", quietly = TRUE))
    install.packages("patchwork")
# "grid" ships with base R itself and never needs installing
```

## 2. Load packages

```{r load-packages, message=FALSE}
# general
library(BiocManager)
library(tidyverse)
library(reshape2)
library(matrixStats)

# network analysis
library(WGCNA)

# statistics
library(lme4)
library(lmerTest)

# functional enrichment
library(topGO)
library(org.Hs.eg.db)

# visualizations
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)
library(corrplot)
library(viridis)
library(cowplot)
library(grid)
library(patchwork)

folder <- "output_coexpression/"
if (!dir.exists(folder)) {
  dir.create(folder)
}
```

## 3. Load and inspect the data

```{r load-data}
# Sample metadata (one row per patient-timepoint)
covid_info_file <- read.csv("input/Sequenced_sample_metadata_020720.csv")
covid_info_file$Pseudotime <- as.character(covid_info_file$Pseudotime)

# Variance-stabilized counts (genes x samples), already normalized upstream
vst_counts <- read.table("output_QC/vst_counts.txt", sep = '\t', header = TRUE)
vst_counts <- vst_counts[, as.character(covid_info_file$RNAseqID)]

# Sample-level metadata, indexed by RNAseqID for easy lookup later
col_data <- covid_info_file
rownames(col_data) <- col_data$RNAseqID
col_data$Gender <- as.factor(col_data$Gender)
col_data$Disease.traj <- as.factor(col_data$Disease.traj)
col_data$Pseudotime <- factor(col_data$Pseudotime)

# Ensembl ID -> gene name/coordinates, so results are human-readable later
ensg2gene <- read.table("input/gencode_v25_gene_info.txt", header = TRUE)
unique_ensg2gene <- subset(ensg2gene, !duplicated(ensg2gene$Gene_id))
rownames(unique_ensg2gene) <- unique_ensg2gene$Gene_id
```

> 🧪 **Exercise (sex balance)**: `table(unique(col_data[, c("PatientID", "Gender")])$Gender)` — how many male vs. female patients are there? Every sex-stratified analysis later in this tutorial (splitting the cohort in two) will have roughly half this sample size per group — worth keeping in mind when interpreting how confidently any "sex difference" result can be trusted.

## 4. Selecting genes for the network

Building a network on every gene is slow and mostly adds noise. We instead keep the genes with the most variable expression across samples, measured by MAD (median absolute deviation) — a version of "spread" that's more robust to outlier samples than plain variance.

```{r gene-variance, fig.width=6, fig.height=4}
vst_counts <- as.matrix(vst_counts)

gene_mad <- apply(vst_counts, 1, mad)

gene_var <- data.frame(Gene_id = rownames(vst_counts),
                        mad = gene_mad,
                        var = rowVars(vst_counts))
gene_var <- merge(gene_var, ensg2gene, by = "Gene_id")

# Rank genes by MAD, most variable first
gene_var <- gene_var[order(gene_var$mad, decreasing = TRUE), ]
gene_var$Seq <- seq_len(nrow(gene_var))

ggplot(gene_var, aes(x = Seq, y = mad)) +
  geom_line() +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  scale_y_log10() +
  labs(x = "Genes (ranked by MAD)", y = "log(expression MAD)") +
  theme_bw()

# Keep the 5000 most variable genes for the network
selected_genes <- gene_var %>%
  arrange(desc(mad)) %>%
  slice_head(n = 5000)

vst_counts_wgcna <- vst_counts[selected_genes$Gene_id, , drop = FALSE]
```

> 🧪 **Exercise**: try `n = 2000` or `n = 10000` instead of 5000 — how does that change the number of modules you get later? More genes isn't automatically better: too many adds noise, too few can miss real structure.

## 5. From counts to a WGCNA-ready matrix

```{r matrix-prep}
# WGCNA expects samples in rows, genes in columns
wgcna_matrix <- t(vst_counts_wgcna)
dim(wgcna_matrix)
```

## 6. Choosing a soft-thresholding power

WGCNA turns raw gene-gene correlations into a weighted network by raising them to a power β. A well-chosen power suppresses weak/noisy correlations while preserving strong ones, and the result should look like a **scale-free network** — a property real biological networks tend to have.

```{r soft-threshold, fig.width=8, fig.height=4}
powers <- c(1:10, seq(12, 30, 2))
sft <- pickSoftThreshold(wgcna_matrix, powerVector = powers, verbose = 0)

par(mfrow = c(1, 2))
plot(sft$fitIndices[, 1], -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft threshold (power)", ylab = "Scale free topology fit, signed R^2",
     type = "n", main = "Scale independence")
text(sft$fitIndices[, 1], -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, col = "red")
abline(h = 0.90, col = "red", lty = 2)

plot(sft$fitIndices[, 1], sft$fitIndices[, 5],
     xlab = "Soft threshold (power)", ylab = "Mean connectivity", type = "n",
     main = "Mean connectivity")
text(sft$fitIndices[, 1], sft$fitIndices[, 5], labels = powers, col = "red")
```

The rule of thumb: pick the **smallest** power where the fit curve (left) crosses ~0.80–0.90, as long as mean connectivity (right) hasn't collapsed to near-zero. Higher powers keep pushing fit up and connectivity down — past a point, that trade is no longer worth it.

```{r pick-power}
soft_power <- sft$powerEstimate
if (is.na(soft_power)) soft_power <- 10  # fallback if pickSoftThreshold can't estimate one
soft_power
```

`powerEstimate` already applies this rule of thumb automatically — trust it over hand-picking a power unless you have a specific reason not to.

> 🧪 **Exercise**: print `sft$fitIndices[, c("Power", "SFT.R.sq", "mean.k.")]` for a couple of powers below and above `soft_power`. Would a different power meaningfully change the connectivity without much change in fit?

> 🧪 **Exercise (foreshadowing Section 19)**: try re-running `pickSoftThreshold()` on just the male samples, and again on just the female samples (`wgcna_matrix[col_data[rownames(wgcna_matrix), "Gender"] == "m", ]`, and similarly for `"f"`). Does the estimated power differ between sexes? If it does, that's already a hint that the two networks may have somewhat different structure before you've even built them — Section 19 reuses one shared `soft_power` for both sexes for comparability, but it's worth knowing whether that's actually the "natural" power for each sex on its own.

## 7. Building the network: adjacency and TOM

```{r adjacency-tom}
adjacency_matrix <- adjacency(wgcna_matrix, power = soft_power, type = "signed")

# Topological Overlap Matrix: two genes are similar not just if they correlate
# directly, but if they share the same neighbors — this makes the network more
# robust to noise than using the raw adjacency alone.
TOM <- TOMsimilarity(adjacency_matrix)
dissTOM <- 1 - TOM
```

## 8. Detecting modules

Modules are found by hierarchical clustering on the dissimilarity, then cutting the tree into branches.

```{r cluster-cut, fig.width=8, fig.height=4}
gene_tree <- hclust(as.dist(dissTOM), method = "average")

module_labels <- cutreeDynamicTree(dendro = gene_tree, minModuleSize = 15, deepSplit = TRUE)
module_colors <- labels2colors(module_labels)

plotDendroAndColors(gene_tree, module_colors, "Module",
                     dendroLabels = FALSE, hang = 0.03,
                     addGuide = TRUE, guideHang = 0.05,
                     main = "Gene clustering and modules")
table(module_colors)
length(unique(module_colors))
```

> 🧪 **Exercise**: try `minModuleSize = 30`. Do you get fewer, larger modules, or does the grey ("unassigned") module just grow? Keep this in mind for Section 19: a smaller `minModuleSize` gives more, smaller modules — and small modules are exactly the ones where a later "sex-specific" call (module preservation, gene overlap) is statistically weakest, since there's less data to test with.

## 9. Checking module similarity before merging

Some modules end up highly correlated with each other — candidates to merge rather than treat as biologically distinct.

```{r module-similarity, fig.width=16, fig.height=8}
# Calculate module eigengenes for the modules found before merging
MEList <- moduleEigengenes(wgcna_matrix, colors = module_colors)
MEs_unmerged <- MEList$eigengenes

# Cluster module eigengenes based on their correlation — modules that cluster
# together are candidates for merging
ME_diss <- 1 - cor(MEs_unmerged)
ME_tree <- hclust(as.dist(ME_diss), method = "average")

# Module pairs joined below this height (i.e. correlation > 1 - cutHeight)
# are the ones mergeCloseModules() will combine
merge_cut_height <- 0.45

ME_cor <- cor(MEs_unmerged)

dendro_grob <- ~{
  plot(ME_tree, main = "Clustering of module eigengenes", xlab = "", sub = "")
  abline(h = merge_cut_height, col = "red", lty = 2)
}

heatmap_grob <- grid.grabExpr(
  draw(Heatmap(ME_cor,
               name = "Correlation",
               col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
               cluster_rows = ME_tree,
               cluster_columns = ME_tree,
               row_names_gp = gpar(fontsize = 9),
               column_names_gp = gpar(fontsize = 9),
               column_title = "Module eigengene similarity (before merging)"))
)

plot_grid(dendro_grob, heatmap_grob, ncol = 2)
```

> 🧪 **Exercise**: any module pair joined below the red dashed line in the dendrogram is what `mergeCloseModules()` will combine next — find one in the heatmap too (a dark red block off the diagonal).

## 10. Merging close modules

```{r merge-modules}
merged <- mergeCloseModules(wgcna_matrix, module_colors, cutHeight = 0.45, verbose = 0)
merged_colors <- merged$colors

table(merged_colors)
```

Compare this to `table(module_colors)` above — that's how many modules got merged.

## 11. Naming and saving modules

```{r save-modules}
gene_module <- data.frame(
  Gene_id     = colnames(wgcna_matrix),
  ModuleColor = merged_colors
)
gene_module <- merge(gene_module, ensg2gene, by = "Gene_id")

# Number modules by size (M1 = largest, excluding the "grey" catch-all module)
module_sizes <- sort(table(gene_module$ModuleColor[gene_module$ModuleColor != "grey"]),
                      decreasing = TRUE)
module_number_map <- setNames(paste0("M", seq_along(module_sizes)), names(module_sizes))
gene_module$Module <- ifelse(gene_module$ModuleColor == "grey", "unassigned",
                              module_number_map[gene_module$ModuleColor])

write.table(gene_module, file.path(folder, "gene_module.txt"),
            quote = FALSE, sep = "\t", row.names = FALSE)
head(gene_module)
```

The **grey** module isn't a real biological module — it's the bucket for genes that didn't fit well anywhere else. Normal to see it, but it's excluded from interpretation from here on.

> 🧪 **Exercise (sex chromosome content)**: even in this one shared, both-sexes network, it's worth a quick sanity check on which modules contain X/Y genes: `table(gene_module$Module, gene_module$Chr %in% c("chrX", "chrY"))`. A module dominated by sex-chromosome genes will almost always show a strong association with sex later — not because it reflects a sex-specific *regulatory program*, but because it's trivially encoding chromosome content. Keep the modules you find here in mind as a comparison point for Section 19, where the same check is done more formally on modules built *separately* per sex.

## 12. Module eigengenes

The eigengene is the first principal component of a module's expression — one number per sample per module.

```{r eigengenes}
MEs0 <- moduleEigengenes(wgcna_matrix, merged_colors)$eigengenes
MEs <- orderMEs(MEs0)
rownames(MEs) <- rownames(wgcna_matrix)

write.table(MEs, file.path(folder, "module_eigengenes.txt"), quote = FALSE, sep = "\t")

MEs_named <- MEs[, colnames(MEs) != "MEgrey"]
colnames(MEs_named) <- module_number_map[gsub("^ME", "", colnames(MEs_named))]
head(MEs_named)
```

## 13. Module behaviour across pseudotime

```{r eigengene-heatmap, fig.width=7, fig.height=5}
pseudotime <- col_data[rownames(MEs_named), "Pseudotime"]

module_by_timepoint <- sapply(split(seq_len(nrow(MEs_named)), pseudotime), function(idx) {
  colMeans(MEs_named[idx, , drop = FALSE], na.rm = TRUE)
})

Heatmap(module_by_timepoint,
        name = "Eigengene",
        col = rev(brewer.pal(11, "RdYlBu")),
        cluster_columns = FALSE,
        column_title = "Pseudotime",
        row_title = "Module")
```

> 🧪 **Exercise**: which module(s) rise early and fall later, versus rise late? What kind of biological process might follow each pattern?

This combined view already hides a question central to this tutorial: does every module follow the *same* trajectory in both sexes, or does the timing/shape of some modules differ by sex? A quick way to check visually, before the formal interaction tests in Sections 16–18:

```{r eigengene-heatmap-by-sex, fig.width=10, fig.height=5}
sex_by_sample <- col_data[rownames(MEs_named), "Gender"]

module_by_timepoint_sex <- function(sex_label) {
  idx_sex <- which(sex_by_sample == sex_label)
  sapply(split(idx_sex, pseudotime[idx_sex]), function(idx) {
    colMeans(MEs_named[idx, , drop = FALSE], na.rm = TRUE)
  })
}

ht_male <- Heatmap(module_by_timepoint_sex("m"),
                    name = "Eigengene (M)", col = rev(brewer.pal(11, "RdYlBu")),
                    cluster_columns = FALSE, column_title = "Male", row_title = "Module")
ht_female <- Heatmap(module_by_timepoint_sex("f"),
                      name = "Eigengene (F)", col = rev(brewer.pal(11, "RdYlBu")),
                      cluster_columns = FALSE, column_title = "Female", row_title = "Module")

ht_male + ht_female
```

> 🧪 **Exercise**: pick a module whose pattern (rising, falling, peaking mid-course) looks visibly different between the two panels above. Hold onto that module — Section 17's systematic interaction test will tell you whether that visual difference is actually statistically supported, or just noise from splitting an already modest sample in half.

## 14. Linking modules to clinical traits

```{r module-trait, fig.width=9, fig.height=6}
clinical_data <- read.csv("input/Clinical_data_020720.csv")
rownames(clinical_data) <- clinical_data$Name

sample_name <- col_data[rownames(MEs_named), "Name"]
common_samples <- intersect(sample_name, rownames(clinical_data))

keep_samples <- sample_name %in% common_samples
ME_by_sample <- MEs_named[keep_samples, ]
clinical_matched <- clinical_data[sample_name[keep_samples], ]
rownames(clinical_matched) <- rownames(ME_by_sample)

# Pseudotime is carried along here too (used later, in Section 17) even
# though it isn't part of the numeric correlation below
clinical_matched <- cbind(clinical_matched,
                           col_data[rownames(ME_by_sample), c("Gender", "Age", "TimepointFromOnset", "Pseudotime")])
clinical_matched$Gender <- as.numeric(factor(clinical_matched$Gender))

# Traits for this Spearman correlation heatmap — all numeric.
# D.Dimers is excluded (too many missing values in this cohort to be reliable).
# Pseudotime is deliberately NOT included here: it's a categorical disease
# stage, not a continuous scale, so it doesn't belong in a correlation matrix
# — it's modeled properly as a factor in Section 17 instead.
trait_vars <- c("CRP", "IL6", "Ferritin", "White.blood.cells", "Lymphocytes",
                 "Gender", "Age", "TimepointFromOnset")
trait_vars <- intersect(trait_vars, colnames(clinical_matched))

module_trait_cor <- cor(ME_by_sample, clinical_matched[, trait_vars], use = "p", method = "spearman")
module_trait_p   <- corPvalueStudent(module_trait_cor, nrow(ME_by_sample))

corrplot::corrplot(module_trait_cor, method = "color", col = viridis::viridis(30),
                    p.mat = module_trait_p, sig.level = c(.001, .01, .05),
                    insig = "label_sig", tl.col = "black", tl.srt = 45)
```

> 🧪 **Exercise**: which module correlates most strongly with CRP? Is it positive or negative?

> 🧪 **Exercise (Gender as a trait)**: `Gender` itself is one of the columns in this heatmap — which module(s), if any, correlate significantly with it? A module correlating with sex here is a *level* difference (its average eigengene differs between males and females) — worth distinguishing from a module whose *relationship to a clinical trait* differs by sex, which is a different question the rest of this tutorial builds toward (Sections 16–18).

## 15. A closer look: a mixed model for one module-trait pair, adjusted for sex

Spearman correlation treats every sample as independent, but several timepoints come from the same patient. A linear mixed model (LMM) with a per-patient random intercept accounts for that — and, from here on, every model also includes `Gender` as a covariate, so any trait-module relationship is estimated *after* accounting for a possible sex-driven shift in the module's baseline level.

```{r module-trait-lmm-single}
model_data <- cbind(ME_by_sample, clinical_matched)
model_data <- model_data[, !duplicated(colnames(model_data))]
model_data$Gender    <- factor(model_data$Gender)
model_data$PatientID <- factor(col_data[rownames(ME_by_sample), "PatientID"])

# Example: does module M3 track with time from symptom onset, adjusting for
# sex, and accounting for repeated timepoints per patient?
fit <- lmer(M3 ~ TimepointFromOnset + Gender + (1 | PatientID), data = model_data)
summary(fit)
```

> 🧪 **Exercise**: change `M3` and `TimepointFromOnset` to a different module/trait pair. Does the estimate change sign? Does it stay significant?

> 🧪 **Exercise (does sex matter here?)**: refit the same model without `Gender` (`lmer(M3 ~ TimepointFromOnset + (1 | PatientID), data = model_data)`). Does the TimepointFromOnset estimate change much? A large change means sex was acting as a confounder for this particular module-trait pair — a small change means it wasn't, for this pair at least. Either way, this is still only a covariate adjustment — it tells you nothing about whether the *slope itself* differs by sex, which is exactly what Section 16 tests next.

## 16. Does the relationship itself differ by sex? (the interaction test)

Adjusting for `Gender` (Section 15) only lets sex shift a module's overall level. It still assumes the *slope* of the module-trait relationship — how much the module eigengene changes per unit of the trait — is the same in both sexes. That assumption is often exactly the interesting thing to question in a sex-sensitive analysis, and it's what a statistical interaction term tests directly.

```{r module-trait-lmm-interaction-single}
# Test whether the TimepointFromOnset effect on module M3 differs by sex, via
# a likelihood ratio test comparing a model with vs. without the interaction
H_1 <- lmer(M3 ~ Gender + TimepointFromOnset + Gender:TimepointFromOnset + (1 | PatientID),
            data = model_data, REML = FALSE)
H_0 <- lmer(M3 ~ Gender + TimepointFromOnset + (1 | PatientID),
            data = model_data, REML = FALSE)

anova(H_0, H_1)
summary(H_1)
```

`anova(H_0, H_1)` tests whether allowing a different slope per sex fits meaningfully better than forcing one shared slope. `summary(H_1)`'s `Gender:TimepointFromOnset` row gives the direction and size of that difference.

```{r plot-associations, fig.width=10, fig.height=5}
p1 <- ggplot(model_data, aes(x = TimepointFromOnset, y = M3)) +
  geom_point(alpha = 0.6, color = "grey40") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(x = "Time from onset", y = "Module M3 eigengene",
       title = "M3 vs. time from onset, sexes combined") +
  theme_bw()

p2 <- ggplot(model_data, aes(x = TimepointFromOnset, y = M3, color = Gender)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Time from onset", y = "Module M3 eigengene",
       title = "M3 vs. time from onset, by sex", color = "Sex") +
  theme_bw()

p1 + p2
```

> 🧪 **Exercise**: repeat with a different module/trait pair from Section 15. Does a visually large gap between the two colored slopes always come with a significant interaction p-value?

> 🧪 **Exercise (why this matters)**: if `M3`'s relationship to `TimepointFromOnset` were reported only from the sexes-combined model (`p1`, left panel) without ever checking the interaction, what conclusion would a reader walk away with — and would it describe both sexes accurately, one sex better than the other, or neither well? This is the core risk of skipping interaction testing in a mixed-sex cohort.

## 17. Generalizing: screening every module x trait pair for sex-specific effects

The single-pair pattern above can be looped across all modules and all traits at once. Pseudotime is added here as a **factor** (a categorical disease stage) — unlike the continuous traits, it isn't `scale()`-d, and a multi-level factor doesn't have a single "estimate," so its overall significance is read off a likelihood ratio test instead of one coefficient.

```{r module-trait-lmm-interaction-loop}
module_names <- colnames(ME_by_sample)

# Same numeric traits as the correlation heatmap (minus Gender, which is
# already a covariate in every model below)
trait_vars_stratified <- setdiff(trait_vars, "Gender")

interaction_results <- expand.grid(Module = module_names, Trait = trait_vars_stratified, stringsAsFactors = FALSE)
interaction_results$Trait_Estimate       <- NA_real_
interaction_results$Trait_P              <- NA_real_
interaction_results$Interaction_Estimate <- NA_real_
interaction_results$Interaction_P_LRT    <- NA_real_

for (i in seq_len(nrow(interaction_results))) {
  module <- interaction_results$Module[i]
  trait  <- interaction_results$Trait[i]

  formula_H1 <- as.formula(paste0(module, " ~ Gender + scale(`", trait, "`) + Gender:scale(`", trait, "`) + (1 | PatientID)"))
  formula_H0 <- as.formula(paste0(module, " ~ Gender + scale(`", trait, "`) + (1 | PatientID)"))

  H_1 <- lmer(formula_H1, data = model_data, REML = FALSE)
  H_0 <- lmer(formula_H0, data = model_data, REML = FALSE)

  # Main trait effect, taken from the no-interaction model (H_0)
  coefs_H0  <- summary(H_0)$coefficients
  trait_row <- grep("^scale\\(", rownames(coefs_H0), value = TRUE)[1]
  interaction_results$Trait_Estimate[i] <- coefs_H0[trait_row, "Estimate"]
  interaction_results$Trait_P[i]        <- coefs_H0[trait_row, "Pr(>|t|)"]

  # Interaction effect, from H_1
  coefs_H1 <- summary(H_1)$coefficients
  interaction_row <- grep(":", rownames(coefs_H1), value = TRUE)[1]

  if (is.na(interaction_row)) {
    message("Skipping interaction for ", module, " x ", trait, ": interaction term dropped (rank-deficient model)")
    next
  }

  lrt <- anova(H_0, H_1)
  interaction_results$Interaction_P_LRT[i]    <- lrt$`Pr(>Chisq)`[2]
  interaction_results$Interaction_Estimate[i] <- coefs_H1[interaction_row, "Estimate"]
}

interaction_results
```

> 🧪 **Exercise**: filter `interaction_results` to rows with `Interaction_P_LRT < 0.05` — how many module-trait pairs show a sex-specific relationship?

> 🧪 **Exercise (a healthy dose of caution)**: this loop tests `length(module_names) * length(trait_vars_stratified)` hypotheses at once, with no multiple-testing correction applied. Roughly how many "significant" hits (p < 0.05) would you expect by chance alone if sex made no real difference to *any* module? Compare that to how many you actually found. This is exactly the kind of check worth doing before presenting a sex-specific finding as real.

## 18. Summarizing trait and interaction effects

```{r plot-all-associations, fig.width=10, fig.height=5}
trait_est <- acast(interaction_results, Module ~ Trait, value.var = "Trait_Estimate")
trait_p   <- acast(interaction_results, Module ~ Trait, value.var = "Trait_P")

interaction_est <- acast(interaction_results, Module ~ Trait, value.var = "Interaction_Estimate")
interaction_p   <- acast(interaction_results, Module ~ Trait, value.var = "Interaction_P_LRT")

redblue <- colorRampPalette(rev(brewer.pal(11, "RdBu")))(200)

par(mfrow = c(1, 2))

corrplot::corrplot(trait_est, method = "color", col = redblue,
                    p.mat = trait_p, sig.level = c(.001, .01, .05),
                    insig = "label_sig", pch.cex = 0.8,
                    tl.col = "black", tl.srt = 45,
                    is.corr = FALSE,
                    title = "Trait effect on module eigengenes",
                    mar = c(0, 0, 2, 0))

corrplot::corrplot(interaction_est, method = "color", col = redblue,
                    p.mat = interaction_p, sig.level = c(.001, .01, .05),
                    insig = "label_sig", pch.cex = 0.8,
                    tl.col = "black", tl.srt = 45,
                    is.corr = FALSE,
                    title = "Sex x trait interaction effect on module eigengenes",
                    mar = c(0, 0, 2, 0))
```

> 🧪 **Exercise**: find a module-trait pair that's significant in the left panel but not the right (a shared effect, not sex-specific) and one that's the other way around (sex-specific, but no strong overall effect). What would each mean biologically?

## 19. Optional advanced extension: comparing networks between sexes

Everything above treats sex as a covariate inside one shared network. A different question: do males and females even have the *same network structure* to begin with? Two ways to check, in increasing order of rigor and runtime:

- **Quick, qualitative check**: build separate networks per sex, then see whether male-defined and female-defined modules contain the same genes (a gene-overlap test). Fast — runs live.
- **Formal module preservation** (`WGCNA::modulePreservation()`): the statistically rigorous way to test this, using permutation testing. Slow (minutes, even at a reduced permutation count) — the rest of this section (from 19.2 onward) is marked `eval=FALSE` and meant to be run afterward, not live.

### 19.1 Building sex-specific networks and checking overlap

```{r sex-specific-networks}
male_samples   <- rownames(wgcna_matrix)[col_data[rownames(wgcna_matrix), "PatientID"] %in%
                                            unique(col_data$PatientID[col_data$Gender == "m"])]
female_samples <- rownames(wgcna_matrix)[col_data[rownames(wgcna_matrix), "PatientID"] %in%
                                            unique(col_data$PatientID[col_data$Gender == "f"])]

wgcna_matrix_male   <- wgcna_matrix[male_samples, ]
wgcna_matrix_female <- wgcna_matrix[female_samples, ]

# Check for genes/samples that are unusable across both sex-specific sets
# (zero variance or excessive missingness in at least one set) before
# building networks, so both sets share the same gene set from the start
multi_expr <- list(Male = list(data = wgcna_matrix_male), Female = list(data = wgcna_matrix_female))
gsg <- goodSamplesGenesMS(multi_expr, verbose = 3)

if (!gsg$allOK) {
  wgcna_matrix_male   <- wgcna_matrix_male[gsg$goodSamples[[1]], gsg$goodGenes]
  wgcna_matrix_female <- wgcna_matrix_female[gsg$goodSamples[[2]], gsg$goodGenes]
}

build_modules <- function(expr_matrix, power) {
  adjacency_matrix <- adjacency(expr_matrix, power = power, type = "signed")
  TOM <- TOMsimilarity(adjacency_matrix)
  dissTOM <- 1 - TOM
  gene_tree <- hclust(as.dist(dissTOM), method = "average")
  module_labels <- cutreeDynamicTree(dendro = gene_tree, minModuleSize = 15, deepSplit = TRUE)
  labels2colors(module_labels)
}

modules_male   <- build_modules(wgcna_matrix_male, soft_power)
modules_female <- build_modules(wgcna_matrix_female, soft_power)

multi_color <- list(Male = modules_male, Female = modules_female)

table(modules_male)
table(modules_female)
```

```{r sex-specific-overlap, fig.width=20, fig.height=16}
# How much do male-defined and female-defined modules overlap, gene for gene?
overlap_table <- table(Male = modules_male, Female = modules_female)

# Fisher's exact test p-value for each male-module x female-module pair
overlap_p <- matrix(NA_real_, nrow = nrow(overlap_table), ncol = ncol(overlap_table),
                     dimnames = dimnames(overlap_table))
total_genes <- length(modules_male)

for (i in rownames(overlap_table)) {
  for (j in colnames(overlap_table)) {
    a <- overlap_table[i, j]
    b <- sum(modules_male == i) - a
    c <- sum(modules_female == j) - a
    d <- total_genes - a - b - c
    overlap_p[i, j] <- fisher.test(matrix(c(a, b, c, d), nrow = 2))$p.value
  }
}

sig_stars <- function(p) {
  if (is.na(p)) return("")
  if (p < 0.001) "***"
  else if (p < 0.01) "**"
  else if (p < 0.05) "*"
  else ""
}

Heatmap(overlap_table,
        name = "Gene\noverlap",
        col = colorRamp2(c(0, max(overlap_table)), c("white", "red")),
        cluster_rows = TRUE, cluster_columns = TRUE,
        row_title = "Male modules", column_title = "Female modules",
        rect_gp = gpar(col = "grey50", lwd = 1),
        cell_fun = function(j, i, x, y, w, h, fill) {
          count <- overlap_table[i, j]
          stars <- sig_stars(overlap_p[i, j])
          grid.text(count, x, y - unit(2, "mm"), gp = gpar(fontsize = 8))
          grid.text(stars, x, y + unit(2, "mm"), gp = gpar(fontsize = 8, col = "black"))
        })
```

```{r sex-specific-modules-overlap}
# A module is "sex-specific" if it doesn't show significant gene overlap
# with any module in the other sex's network (excluding "grey")
male_modules_named   <- setdiff(rownames(overlap_table), "grey")
female_modules_named <- setdiff(colnames(overlap_table), "grey")

overlap_p_named <- overlap_p[male_modules_named, female_modules_named, drop = FALSE]

male_specific   <- male_modules_named[apply(overlap_p_named, 1, function(p) all(p >= 0.05, na.rm = TRUE))]
female_specific <- female_modules_named[apply(overlap_p_named, 2, function(p) all(p >= 0.05, na.rm = TRUE))]

male_specific
female_specific
```

> 🧪 **Exercise**: finding no sex-specific modules here is a real, useful result too — it means network *membership* looks similar between sexes, even if (as Section 18 showed) a shared module's *relationship to a trait* can still differ by sex.

### 19.2 Formal module preservation (slow)

```{r module-preservation, eval=FALSE}
# ------------------------------------------------------------------
# modulePreservation() formally tests whether modules found in one sex
# ("reference") are reproducible in the other ("test"), using a
# permutation-based Z-summary statistic (Langfelder et al. 2011).
# Z-summary > 10 = strong preservation, 2-10 = weak/moderate, < 2 = not
# preserved. Much more rigorous than the overlap heatmap above, but
# permutation testing is slow — expect several minutes even here.
# ------------------------------------------------------------------

multi_expr <- list(Male = list(data = wgcna_matrix_male), Female = list(data = wgcna_matrix_female))

preservation <- modulePreservation(
  multiData = multi_expr,
  multiColor = multi_color,
  referenceNetworks = 1:2,      # test each sex as the reference in turn
  nPermutations = 50,           # reduced from the default 200 for speed
  randomSeed = 1,
  verbose = 1
)

male_ref_summary   <- preservation$preservation$Z$ref.Male$inColumnsAlsoPresentIn.Female
female_ref_summary <- preservation$preservation$Z$ref.Female$inColumnsAlsoPresentIn.Male

male_ref_summary[order(-male_ref_summary$Zsummary.pres), c("moduleSize", "Zsummary.pres")]
female_ref_summary[order(-female_ref_summary$Zsummary.pres), c("moduleSize", "Zsummary.pres")]
```

```{r sex-specific-modules-preservation, eval=FALSE}
# Zsummary.pres < 2 = not preserved -> treated as sex-specific here.
# "gold" is WGCNA's internal random-gene reference module, not a real one —
# excluded along with "grey" (unassigned genes).
exclude_modules <- c("grey", "gold")

male_specific_preservation <- rownames(male_ref_summary)[
  !(rownames(male_ref_summary) %in% exclude_modules) & male_ref_summary$Zsummary.pres < 2
]

female_specific_preservation <- rownames(female_ref_summary)[
  !(rownames(female_ref_summary) %in% exclude_modules) & female_ref_summary$Zsummary.pres < 2
]

male_specific_preservation
female_specific_preservation
```

`Zsummary.pres` is sensitive to module size — small modules score lower even when genuinely preserved, since the permutation test has less power with fewer genes. Treat a small module more cautiously than a large one.

```{r sex-specific-modules-preservation-genes, eval=FALSE}
male_specific_genes_pres <- lapply(male_specific_preservation, function(mod) {
  genes <- colnames(wgcna_matrix_male)[modules_male == mod]
  unique_ensg2gene[genes, ]
})
names(male_specific_genes_pres) <- male_specific_preservation

female_specific_genes_pres <- lapply(female_specific_preservation, function(mod) {
  genes <- colnames(wgcna_matrix_female)[modules_female == mod]
  unique_ensg2gene[genes, ]
})
names(female_specific_genes_pres) <- female_specific_preservation

male_specific_genes_pres
female_specific_genes_pres
```

```{r module-size-table, eval=FALSE}
male_size_table <- data.frame(
  Sex = "Male",
  Module = names(male_specific_genes_pres),
  n_genes = sapply(male_specific_genes_pres, nrow)
)

female_size_table <- data.frame(
  Sex = "Female",
  Module = names(female_specific_genes_pres),
  n_genes = sapply(female_specific_genes_pres, nrow)
)

module_size_table <- rbind(male_size_table, female_size_table)
module_size_table
```

Modules noticeably smaller than the rest are the ones to treat with the most caution (see the note above on `Zsummary.pres` and size).

```{r sex-chromosome-check, eval=FALSE}
# Before reading anything biological into a "sex-specific" module, rule out
# the boring explanation: is it just sitting on a sex chromosome?
summarize_sex_chromosomes <- function(gene_df_list) {
  do.call(rbind, lapply(names(gene_df_list), function(mod) {
    chr <- gene_df_list[[mod]]$Chr
    data.frame(
      Module        = mod,
      n_genes       = length(chr),
      n_X           = sum(chr == "chrX", na.rm = TRUE),
      n_Y           = sum(chr == "chrY", na.rm = TRUE),
      pct_sex_chrom = round(100 * sum(chr %in% c("chrX", "chrY"), na.rm = TRUE) / length(chr), 1)
    )
  }))
}

male_chrom_summary   <- summarize_sex_chromosomes(male_specific_genes_pres)
male_chrom_summary$Sex <- "Male"

female_chrom_summary <- summarize_sex_chromosomes(female_specific_genes_pres)
female_chrom_summary$Sex <- "Female"

chrom_summary <- rbind(male_chrom_summary, female_chrom_summary)
chrom_summary[, c("Sex", "Module", "n_genes", "n_X", "n_Y", "pct_sex_chrom")]
```

A high `pct_sex_chrom` means the module is "sex-specific" mainly because of its X/Y content — a much less interesting story than a sex-specific *autosomal* program, which is what to prioritize below.

> If, when you run this, a Y-chromosome gene shows up in a module you called "female," check whether it's uniformly low across all female samples (likely a network-construction artifact/cross-mapping noise) or clearly elevated in a specific sample (worth cross-checking that sample's sex label before trusting any downstream result from it).

### 19.3 A worked example: following up on one module

The rest of this section walks through one module in depth — substitute whichever module (and sex) your own run flags as the most interesting, largest, and most clearly autosomal.

```{r module-eigengene-by-sex, eval=FALSE}
# Replace "black" / wgcna_matrix_female / modules_female below with whichever
# module and sex your own preservation results point to
module_of_interest <- "black"
module_genes <- colnames(wgcna_matrix_female)[modules_female == module_of_interest]

# Eigengene computed once, on the combined matrix, so both sexes are on the
# same scale for direct comparison
module_eigengene <- moduleEigengenes(wgcna_matrix[, module_genes],
                                      colors = rep(module_of_interest, length(module_genes)))$eigengenes

module_df <- data.frame(
  Sample = rownames(wgcna_matrix),
  ME = module_eigengene[[paste0("ME", module_of_interest)]],
  Gender = col_data[rownames(wgcna_matrix), "Gender"]
)

ggplot(module_df, aes(x = Gender, y = ME, fill = Gender)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  labs(y = paste(module_of_interest, "module eigengene"), title = "Module eigengene, by sex") +
  theme_bw()
```

This eigengene-by-sex plot alone won't necessarily show a mean-level difference — sex-specificity here is about *network structure*, not eigengene magnitude. The next chunk checks that more directly.

```{r module-connectivity-comparison, eval=FALSE}
adj_female <- adjacency(wgcna_matrix_female[, module_genes], power = soft_power, type = "signed")
adj_male   <- adjacency(wgcna_matrix_male[, module_genes], power = soft_power, type = "signed")

conn_female <- rowSums(adj_female) - 1  # subtract self-connection
conn_male   <- rowSums(adj_male) - 1

conn_compare <- data.frame(Gene = module_genes, Female = conn_female, Male = conn_male)

ggplot(conn_compare, aes(x = Male, y = Female)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  labs(title = paste(module_of_interest, "module: gene connectivity, female vs. male network"),
       x = "Connectivity in male network", y = "Connectivity in female network") +
  theme_bw()
```

Genes far above the diagonal (highly connected in one sex's network, not the other) are the ones actually driving the "not preserved" call — this is the plot that most directly matches what `modulePreservation()` tested.

```{r module-clinical-relevance, eval=FALSE}
module_df$Name <- col_data[rownames(wgcna_matrix), "Name"]
module_df_female <- module_df[module_df$Gender == "f", ]

module_clinical <- clinical_matched[module_df_female$Sample, trait_vars_stratified]
module_clinical$ME <- module_df_female$ME
module_clinical$PatientID <- col_data[rownames(wgcna_matrix)[match(module_df_female$Name, col_data$Name)], "PatientID"]

# Continuous traits get a shared scatter grid; Pseudotime (categorical) gets
# its own boxplot rather than sharing an x-axis with numeric traits
continuous_traits <- setdiff(trait_vars_stratified, "Pseudotime")
module_clinical_long <- melt(module_clinical[, c("ME", "PatientID", continuous_traits)],
                              id.vars = c("ME", "PatientID"),
                              variable.name = "Trait", value.name = "Value")

ggplot(module_clinical_long, aes(x = Value, y = ME)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  facet_wrap(~ Trait, scales = "free_x") +
  labs(y = paste(module_of_interest, "module eigengene"), x = NULL,
       title = "Module eigengene vs. clinical traits (females only)") +
  theme_bw()

```

```{r module-lmm-clinical, eval=FALSE}
module_lmm_results <- data.frame(Trait = trait_vars_stratified,
                                  Estimate = NA_real_, P.value = NA_real_)

for (i in seq_along(trait_vars_stratified)) {
  trait <- trait_vars_stratified[i]
  trait_term <- if (trait == "Pseudotime") "Pseudotime" else paste0("scale(`", trait, "`)")
  formula_i <- as.formula(paste0("ME ~ ", trait_term, " + (1 | PatientID)"))
  fit <- lmer(formula_i, data = module_clinical)

  if (trait == "Pseudotime") {
    fit_null <- lmer(ME ~ 1 + (1 | PatientID), data = module_clinical)
    module_lmm_results$P.value[i] <- anova(fit_null, fit)$`Pr(>Chisq)`[2]
    # Estimate stays NA — no single coefficient for a multi-level factor
  } else {
    coefs <- summary(fit)$coefficients
    trait_row <- grep("^scale\\(", rownames(coefs), value = TRUE)[1]
    module_lmm_results$Estimate[i] <- coefs[trait_row, "Estimate"]
    module_lmm_results$P.value[i]  <- coefs[trait_row, "Pr(>|t|)"]
  }
}

module_lmm_results[order(module_lmm_results$P.value), ]
```

```{r module-trajectory-plot, eval=FALSE}
ggplot(module_clinical, aes(x = TimepointFromOnset, y = ME, color = CRP)) +
  geom_point(size = 2) +
  scale_color_viridis_c() +
  geom_smooth(method = "loess", se = FALSE, color = "black", linetype = "dashed") +
  labs(y = paste(module_of_interest, "module eigengene"), x = "Time from onset",
       title = "Module trajectory (females), colored by CRP") +
  theme_bw()
```

If this module shows a real, significant relationship to a clinical trait *within one sex* — a relationship you couldn't even test the same way in the other sex, since the network structure doesn't reproduce there — that ties together the whole thread: soft-thresholding → modules → preservation → connectivity confirmation → clinical relevance. One honest caveat: this is one module tested against several traits in a fairly small subset of patients, and without correction for multiple testing — treat a hit here as a hypothesis worth validating in a larger cohort, not a confirmed finding.

### 19.4 Functional enrichment of the module

```{r module-go-enrichment, eval=FALSE}
library(topGO)
library(org.Hs.eg.db)

# Universe = every gene that went into the network — the appropriate
# background for "what's over-represented in this module relative to the
# genes we actually tested," not the whole genome
gene_universe <- colnames(wgcna_matrix)

geneList <- factor(as.integer(gene_universe %in% module_genes))
names(geneList) <- gene_universe

GOdata <- new("topGOdata",
              ontology = "BP",
              allGenes = geneList,
              geneSel = function(x) x == 1,
              annot = annFUN.org,
              mapping = "org.Hs.eg.db",
              ID = "ensembl",
              nodeSize = 5)  # ignore GO terms with fewer than 5 annotated genes

result_fisher <- runTest(GOdata, algorithm = "classic", statistic = "fisher")

go_results <- GenTable(GOdata, classicFisher = result_fisher,
                        orderBy = "classicFisher", topNodes = 20)
go_results$classicFisher <- as.numeric(gsub("< ", "", go_results$classicFisher))

go_results
```

```{r module-go-enrichment-plot, eval=FALSE}
go_results$Term <- factor(go_results$Term, levels = rev(go_results$Term))

ggplot(go_results[1:15, ], aes(x = Term, y = -log10(classicFisher))) +
  geom_col(fill = "black") +
  coord_flip() +
  labs(x = NULL, y = "-log10(p-value)",
       title = paste("Top enriched GO terms —", module_of_interest, "module")) +
  theme_bw()
```
