---
title: "Single-cell transcriptomics with a sex & gender lens"
subtitle: "Part 2 - sex-sensitive downstream analyses"
author: "Joana P. Bernardes & Neha Mishra"
date: "MOSAIC-ID workshop 2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
params:
  input_rds: "output/D3_annotated_6samples.rds"  # written at the end of Part 1
  out_dir: "output"
  celltype_col: "CellTypes"
  celltype_of_interest: ""   # empty = choose automatically from the differential abundance results
  min_cells: 10              # minimum cells per sample for a pseudobulk sample
  solutions: false           # TRUE renders the instructor version (with solutions)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(33)
```

# Introduction

In Part 1 we annotated about a dozen cell types in 6 PBMC samples. Now we ask
biological questions. With a sex lens, "Is there a difference?" splits into
**three different questions**, and each needs its own statistical model:

| Question | Statistical term | Model |
|---|---|---|
| Does COVID-19 change X, on average over both sexes? | main effect of disease (adjusted for sex) | `~ sex + disease` |
| Do female and male samples differ in X? | main effect of sex (adjusted for disease) | `~ sex + disease` |
| Is the effect of COVID-19 **different** in females and males? | sex × disease **interaction** | `~ sex * disease` |

Four ideas come back throughout this notebook:

1. **The unit of replication is the individual, not the cell.** 5,000 cells
   from one woman are still *one* woman. Treating cells as replicates
   (*pseudoreplication*) gives tiny, misleading p-values.
2. **Sex-stratified results are useful, but they do not prove a sex
   difference.** "Significant in females, not significant in males" is not a
   test of a sex difference (Gelman & Stern, 2006). The interaction test is.
3. **Sex-chromosome genes** (*XIST*, chrY genes, X-inactivation escapees)
   always differ between the sexes. They can hide, or fake, other results if the
   design is unbalanced.
4. **Power.** With one or two donors per group, a non-significant result
   tells us very little. We are *exploring* and *learning the methods* here,
   not making claims about COVID-19 biology.

Outline:

1. Differential abundance (DA): do cell-type proportions change?
2. Differential expression (DE), per cell vs. pseudobulk, with sex in the model
3. Enrichment analysis: over-representation (ORA) and gene set enrichment
   (GSEA), by sex

# 0. Setup

```{r libraries}
library(Seurat)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(limma)
library(edgeR)
library(speckle)       # propeller (differential abundance)
library(DESeq2)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)

# Bioconductor packages mask some dplyr verbs (e.g. AnnotationDbi::select, matrixStats::count)
select <- dplyr::select; count <- dplyr::count; filter <- dplyr::filter

dir.create(params$out_dir, showWarnings = FALSE, recursive = TRUE)
sex_cols     <- c(female = "#E69F00", male = "#009E73")  # colour-blind safe, avoids pink/blue stereotypes
disease_cols <- c(healthy = "grey60", COVID19 = "#CC3311")
theme_set(theme_bw())
```

# 1. Data and study design

```{r load}
obj <- readRDS(params$input_rds)
obj$CellTypes <- as.character(obj[[params$celltype_col, drop = TRUE]])
obj$disease   <- factor(obj$disease, levels = c("healthy", "COVID19"))  # healthy = reference level
obj$sex       <- factor(obj$sex,     levels = c("female", "male"))       # female  = reference level
obj$group     <- factor(paste(obj$sex, obj$disease, sep = "_"),
                        levels = c("female_healthy", "male_healthy", "female_COVID19", "male_COVID19"))
Idents(obj) <- "CellTypes"
obj

# one row per sample
sample_info <- obj@meta.data %>%
  distinct(sample_id, donor, disease, sex, group, timepoint) %>%
  arrange(disease, sex, sample_id)
rownames(sample_info) <- sample_info$sample_id
sample_info

# chromosome of each gene (computed in Part 1; recomputed if missing)
gene_chr <- obj@misc$gene_chr
if (is.null(gene_chr)) {
  valid <- intersect(rownames(obj), keys(org.Hs.eg.db, keytype = "SYMBOL"))
  gene_chr <- AnnotationDbi::mapIds(org.Hs.eg.db, keys = valid, keytype = "SYMBOL", column = "CHR",
                                    multiVals = function(x) paste(sort(unique(x)), collapse = ";"))
}
gene_chr <- setNames(ifelse(is.na(gene_chr[rownames(obj)]), "unknown", gene_chr[rownames(obj)]), rownames(obj))
chr_class <- function(genes) {
  ch <- gene_chr[genes]
  factor(case_when(ch == "X" ~ "chrX", ch %in% c("Y", "X;Y") ~ "chrY / PAR", TRUE ~ "autosomal / other"),
         levels = c("chrX", "chrY / PAR", "autosomal / other"))
}
```

```{r design-table}
table(sample_info$disease, sample_info$sex)                                    # samples
table(distinct(sample_info, donor, disease, sex)[, c("disease", "sex")])       # donors
repeated_donors <- any(duplicated(sample_info$donor))
repeated_donors
```

**Our design:** 4 donors, one per sex × disease group. Each COVID-19 donor
was sampled at two time points (T1, T2), so we have 6 samples.

| | female | male |
|---|---|---|
| COVID-19 | donor COV_F: T1, T2 | donor COV_M: T1, T2 |
| healthy | donor HC_F | donor HC_M |

> **Exercise 1 (5 min, pen & paper).**
>
> 1. How many biological replicates (donors) are there per sex × disease
>    group?
> 2. A linear model with `~ sex + disease` estimates 3 parameters
>    (intercept, sex, disease). If we treat the 6 samples as independent,
>    how many residual degrees of freedom are left? And if we count
>    donors (4) instead of samples?
> 3. The two COVID-19 samples of a sex come from the **same person**. What
>    is the variability between them actually measuring?
> 4. The "sex difference in healthy people" is estimated from how many
>    women and how many men? And the sex × disease interaction?

```{asis, echo=params$solutions}
**Solution.** (1) One donor per group. (2) With samples: 6 - 3 = 3 residual
df (6 - 4 = 2 for `~ sex * disease`). With donors: 4 - 3 = 1, and 0 for the
interaction model, so nothing is left to estimate the between-person
variability. (3) T1 vs T2 of one patient measures *within-person* variability
(disease course and technical noise), not *between-person* variability. When
we treat the time points as replicates, the COVID-19 groups look less variable
than people really are, so p-values are too small. The healthy groups have no
replicate at all. (4) One woman vs one man. The interaction compares
(COV_F − HC_F) with (COV_M − HC_M), i.e. four individuals, so any individual
difference (age, genetics, severity, day of sampling) is inseparable from
"sex". **Everything we compute today is a demonstration of the methods, not
evidence of sex differences in COVID-19.** We come back to this in 2.4, 3.6
and 5.
```

# 2. Differential abundance: cell proportions

## 2.1 Look at the data first

```{r proportions}
ct_counts <- table(sample_id = obj$sample_id, celltype = obj$CellTypes)
ct_counts

prop_df <- as.data.frame(prop.table(ct_counts, margin = 1), responseName = "proportion") %>%
  left_join(sample_info, by = "sample_id")
```

```{r prop-barplot, fig.width=10, fig.height=5}
ggplot(prop_df, aes(x = sample_id, y = proportion, fill = celltype)) +
  geom_col() +
  facet_grid(~ disease + sex, scales = "free_x", space = "free_x") +
  labs(x = NULL, y = "fraction of cells", title = "Cell-type composition per sample") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r prop-dotplot, fig.width=12, fig.height=8}
ggplot(prop_df, aes(x = disease, y = 100 * proportion, colour = sex)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  stat_summary(aes(group = sex), fun = mean, geom = "line", position = position_dodge(width = 0.5)) +
  facet_wrap(~ celltype, scales = "free_y") +
  scale_colour_manual(values = sex_cols) +
  labs(x = NULL, y = "% of PBMCs", title = "Every dot is one sample")
```

> **Exercise 2.1.** Which cell types look different between COVID-19 and
> healthy? Is the change in the same direction in female and male samples?
> Are the lines parallel (no interaction) or not (possible interaction)?

## 2.2 The naive way (do not do this!)

A tempting approach is to pool all cells per group and run a chi-squared test
per cell type:

```{r naive-da}
naive <- lapply(colnames(ct_counts), function(ct) {
  in_ct <- obj$CellTypes == ct
  data.frame(celltype = ct,
             p_disease = chisq.test(table(in_ct, obj$disease))$p.value,
             p_sex     = chisq.test(table(in_ct, obj$sex))$p.value)
}) %>% bind_rows() %>% mutate(across(starts_with("p_"), ~ signif(.x, 2)))
naive
```

> **Exercise 2.2 (discussion).** How many cell types are "significant"?
> What does the test assume about the cells? Suppose one healthy woman
> simply has more NK cells than average. Would this test notice that this is
> about *one person*?

```{asis, echo=params$solutions}
**Solution.** Almost everything comes out significant because the test treats
every cell as an independent observation (n = thousands). Cells from the same
person are not independent. The test ignores the (large) variation *between
people*, so it tests "are these particular 6 samples different?" instead of
"are COVID-19 patients different from healthy people?". This is
**pseudoreplication**.
```

## 2.3 Sample-level test: propeller

`propeller` (Phipson *et al.*, 2022, *Bioinformatics*):

1. Computes the proportion of each cell type **per sample**.
2. Transforms the proportions (logit or arcsin-square-root).
3. Fits a linear model per cell type with limma. Each sample is one
   observation, so we can include `sex` in the design.

```{r propeller}
props <- getTransformedProps(clusters = obj$CellTypes, sample = obj$sample_id, transform = "logit")
si <- sample_info[colnames(props$TransformedProps), ]   # same sample order as the proportions
# propeller wants a "group means" design (no intercept) plus covariates, and a contrast
# between two group columns. Both designs below are the same model, ~ disease + sex;
# they only differ in which variable is written as the group means.
design_disease <- model.matrix(~ 0 + disease + sex, data = si)
colnames(design_disease)        # diseasehealthy, diseaseCOVID19, sexmale
da_disease <- propeller.ttest(props, design = design_disease, contrasts = c(-1, 1, 0),  # COVID19 - healthy
                              robust = TRUE, trend = FALSE, sort = FALSE)

design_sex <- model.matrix(~ 0 + sex + disease, data = si)
colnames(design_sex)            # sexfemale, sexmale, diseaseCOVID19
da_sex <- propeller.ttest(props, design = design_sex, contrasts = c(-1, 1, 0),          # male - female
                          robust = TRUE, trend = FALSE, sort = FALSE)
da_disease
da_sex
```

```{r da-summary}
mean_prop <- prop_df %>% group_by(celltype) %>%
  summarise(pct_healthy = 100 * mean(proportion[disease == "healthy"]),
            pct_COVID19 = 100 * mean(proportion[disease == "COVID19"]),
            pct_female  = 100 * mean(proportion[sex == "female"]),
            pct_male    = 100 * mean(proportion[sex == "male"]),
            min_cells   = min(ct_counts[, as.character(celltype[1])]))

da <- data.frame(celltype    = rownames(da_disease),
                 t_disease   = da_disease$Tstatistic, p_disease = da_disease$P.Value, fdr_disease = da_disease$FDR,
                 t_sex       = da_sex[rownames(da_disease), "Tstatistic"],
                 p_sex       = da_sex[rownames(da_disease), "P.Value"],
                 fdr_sex     = da_sex[rownames(da_disease), "FDR"]) %>%
  left_join(mean_prop, by = "celltype") %>%
  arrange(pmax(p_disease, p_sex))
da %>% mutate(across(where(is.double), ~ signif(.x, 3)))
write.csv(da, file.path(params$out_dir, "DA_propeller_disease_sex.csv"), row.names = FALSE)
```

> **Exercise 2.3.** Compare the propeller p-values with the naive ones.
> Which cell types stay significant? Why are the p-values so much larger?
> A positive `t_sex` means higher in male samples, a positive `t_disease`
> higher in COVID-19. Which cell types differ by sex?

## 2.4 Going further (optional)

> **Exercise 2.4 (interaction).** Fit `~ sex * disease` and test the
> interaction coefficient (`sexmale:diseaseCOVID19`). Is the COVID-19 effect
> on any cell type different between the sexes? How many residual degrees of
> freedom does this model have?

```{r ex2.4-solution, eval=params$solutions, include=params$solutions}
# ~ sex * disease has the same fit as one mean per sex x disease group
design_int <- model.matrix(~ 0 + group, data = si)
colnames(design_int)   # female_healthy, male_healthy, female_COVID19, male_COVID19
# (female_COVID19 - female_healthy) - (male_COVID19 - male_healthy)
da_int <- propeller.ttest(props, design = design_int, contrasts = c(-1, 1, 1, -1),
                          robust = TRUE, trend = FALSE, sort = TRUE)
da_int
# The same test with limma and the classical interaction coding:
fit_int <- eBayes(lmFit(props$TransformedProps, model.matrix(~ sex * disease, data = si)), robust = TRUE)
topTable(fit_int, coef = "sexmale:diseaseCOVID19", sort.by = "P")
# residual df = 6 samples - 4 coefficients = 2 -> very little power: absence of evidence != evidence of absence
```

> **Exercise 2.5 (repeated donors).** Our COVID-19 donors contributed two
> samples each, and these are not independent. Two ways to handle this:
>
> 1. limma's `duplicateCorrelation()` with `block = donor`. Try it on
>    `props$TransformedProps`. What happens to the p-values?
> 2. Average the two time points per donor and re-run propeller on the 4
>    donors. How many residual degrees of freedom are left?

```{r ex2.5-solution, eval=params$solutions && repeated_donors, include=params$solutions && repeated_donors}
design_add <- model.matrix(~ sex + disease, data = si)
dc  <- duplicateCorrelation(props$TransformedProps, design_add, block = si$donor)
dc$consensus.correlation   # correlation between samples of the same donor
fit <- lmFit(props$TransformedProps, design_add, block = si$donor, correlation = dc$consensus.correlation)
fit <- eBayes(fit, robust = TRUE)
topTable(fit, coef = "diseaseCOVID19", sort.by = "P")
topTable(fit, coef = "sexmale",        sort.by = "P")
# The correlation is estimated from only 2 donors with repeated samples, so it is itself very uncertain.

# (2) donor level: sum the cells of both time points per donor
props_donor <- getTransformedProps(clusters = obj$CellTypes, sample = obj$donor, transform = "logit")
si_donor <- distinct(sample_info, donor, sex, disease)
rownames(si_donor) <- si_donor$donor
si_donor <- si_donor[colnames(props_donor$TransformedProps), ]
design_donor <- model.matrix(~ 0 + disease + sex, data = si_donor)
nrow(design_donor) - ncol(design_donor)     # residual df = 1
propeller.ttest(props_donor, design = design_donor, contrasts = c(-1, 1, 0),
                robust = TRUE, trend = FALSE, sort = TRUE)
# With 1 residual df the p-values rest almost entirely on the prior borrowed across cell types.
```

Other DA methods you may meet: `sccomp` (Bayesian compositional model),
`scCODA` (Python), and neighbourhood-based methods such as `miloR`, which
do not need clusters.

## 2.5 Choose the cell type for the downstream analyses

We look for a cell type that changes with **disease** and differs by
**sex**, and that has at least `min_cells` cells in **every** sample (needed
for pseudobulk). Rule, from strict to lenient:

1. FDR < 0.05 for both disease and sex
2. otherwise nominal p < 0.05 for both
3. otherwise the cell type with the smallest max(p_disease, p_sex)

Instructors can override the choice with `params$celltype_of_interest`.

**Caveat for our data:** these p-values treat the two time points of each
COVID-19 donor as independent replicates, so they are optimistic (see
Exercise 2.5). We use them only to *pick an example*, not to make claims.

```{r choose-celltype}
candidates <- da %>% filter(min_cells >= params$min_cells)
rule <- if (any(candidates$fdr_disease < 0.05 & candidates$fdr_sex < 0.05)) {
  candidates <- filter(candidates, fdr_disease < 0.05 & fdr_sex < 0.05); "FDR < 0.05 for disease and sex"
} else if (any(candidates$p_disease < 0.05 & candidates$p_sex < 0.05)) {
  candidates <- filter(candidates, p_disease < 0.05 & p_sex < 0.05);   "nominal p < 0.05 for disease and sex"
} else "no cell type passes p < 0.05 for both - taking the smallest max(p) (exploratory!)"

coi <- if (nzchar(params$celltype_of_interest)) params$celltype_of_interest else
  candidates$celltype[which.min(pmax(candidates$p_disease, candidates$p_sex))]
if (nzchar(params$celltype_of_interest)) rule <- "set by the instructor (params$celltype_of_interest)"
stopifnot(coi %in% obj$CellTypes)

cat("Cell type of interest:", coi, "\nSelection rule:", rule, "\n")
da %>% filter(celltype == coi)
```

```{r coi-plot, fig.width=5, fig.height=4}
ggplot(filter(prop_df, celltype == coi), aes(x = disease, y = 100 * proportion, colour = sex)) +
  geom_point(size = 4, position = position_dodge(width = 0.4)) +
  scale_colour_manual(values = sex_cols) +
  labs(x = NULL, y = paste("%", coi, "of PBMCs"), title = coi)
```

# 3. Differential expression in the cell type of interest

```{r subset}
sub <- obj[, obj$CellTypes == coi]
table(sub$sample_id, sub$group)
```

## 3.1 Per-cell DE (as in many papers - and in our 2022 course)

`FindMarkers()` compares **cells**: every cell is treated as a replicate.

```{r cell-level-de}
de_cell_disease <- FindMarkers(sub, group.by = "disease", ident.1 = "COVID19", ident.2 = "healthy",
                               test.use = "wilcox", logfc.threshold = 0.1, verbose = FALSE)
sum(de_cell_disease$p_val_adj < 0.05)   # number of "significant" genes
head(de_cell_disease, 10)
```

> **Exercise 3.1 (method comparison, from the 2022 course).** Repeat the
> test with `test.use = "bimod"` (and `"MAST"`, if installed). How many
> genes are significant with each method, and how many are shared?

```{r ex3.1-solution, eval=params$solutions, include=params$solutions}
de_bimod <- FindMarkers(sub, group.by = "disease", ident.1 = "COVID19", ident.2 = "healthy",
                        test.use = "bimod", logfc.threshold = 0.1, verbose = FALSE)
wil   <- rownames(de_cell_disease)[de_cell_disease$p_val_adj < 0.05]
bimod <- rownames(de_bimod)[de_bimod$p_val_adj < 0.05]
c(wilcox = length(wil), bimod = length(bimod), shared = length(intersect(wil, bimod)))
if (requireNamespace("MAST", quietly = TRUE)) {
  de_mast <- FindMarkers(sub, group.by = "disease", ident.1 = "COVID19", ident.2 = "healthy",
                         test.use = "MAST", logfc.threshold = 0.1, verbose = FALSE)
  mast <- rownames(de_mast)[de_mast$p_val_adj < 0.05]
  c(MAST = length(mast), shared_all = length(Reduce(intersect, list(wil, bimod, mast))))
}
```

## 3.2 Sex differences per cell: meet the sex-chromosome genes

```{r cell-level-sex, fig.width=7, fig.height=5}
de_cell_sex <- FindMarkers(sub, group.by = "sex", ident.1 = "male", ident.2 = "female",
                           test.use = "wilcox", logfc.threshold = 0.1, verbose = FALSE)
de_cell_sex$gene <- rownames(de_cell_sex)
de_cell_sex$chr  <- chr_class(de_cell_sex$gene)

top_sex <- de_cell_sex %>% filter(p_val_adj < 0.05) %>% slice_max(abs(avg_log2FC), n = 15)
ggplot(de_cell_sex, aes(avg_log2FC, -log10(p_val_adj + 1e-300), colour = chr)) +
  geom_point(size = 1) +
  geom_text_repel(data = top_sex, aes(label = gene), size = 3, max.overlaps = 30, show.legend = FALSE) +
  scale_colour_manual(values = c("chrX" = "#E69F00", "chrY / PAR" = "#009E73", "autosomal / other" = "grey70")) +
  labs(x = "log2FC (male vs female)", y = "-log10 adj. p", title = paste(coi, "- male vs female, per cell"))

de_cell_sex %>% filter(p_val_adj < 0.05) %>% count(chr) %>% mutate(pct = round(100 * n / sum(n), 1))
```

> **Exercise 3.2.**
>
> 1. What are the strongest sex-associated genes? On which chromosomes are
>    they?
> 2. Many chrY genes have an X-linked partner (a *gametolog*): *DDX3X/DDX3Y*,
>    *KDM5C/KDM5D*, *KDM6A/UTY*, *EIF1AX/EIF1AY*, *ZFX/ZFY*, *RPS4X/RPS4Y1*.
>    Look them up in `de_cell_sex`. Why could the X copy be higher in
>    females? (Hint: X-inactivation escape; Tukiainen *et al.*, 2017,
>    *Nature*.)
> 3. Now repeat the male vs female test using **only the two healthy
>    samples** (one woman, one man). How many genes are "significant"? Would
>    you publish these as "sex differences in `r coi`"?

```{r ex3.2-solution, eval=params$solutions, include=params$solutions}
gametologs <- c("DDX3X", "DDX3Y", "KDM5C", "KDM5D", "KDM6A", "UTY", "EIF1AX", "EIF1AY", "ZFX", "ZFY", "RPS4X", "RPS4Y1")
de_cell_sex[intersect(gametologs, rownames(de_cell_sex)), c("avg_log2FC", "pct.1", "pct.2", "p_val_adj", "chr")]

sub_hc <- sub[, sub$disease == "healthy"]
de_hc_sex <- FindMarkers(sub_hc, group.by = "sex", ident.1 = "male", ident.2 = "female",
                         logfc.threshold = 0.1, verbose = FALSE)
sum(de_hc_sex$p_val_adj < 0.05)
# Many "significant" genes from n = 1 vs n = 1! These are differences between two PEOPLE
# (genetics, age, time of sampling, processing batch...), not between the sexes.
```

## 3.3 Pseudobulk: one expression profile per sample

We **sum the raw counts of all cells of the cell type in each sample**. Each
sample then becomes one "bulk RNA-seq" library, and we can use the bulk tools
from Day 2 (DESeq2, edgeR, limma). They treat samples, not cells, as
replicates, and let us put sex and disease in the design.

```{r pseudobulk}
pseudobulk <- function(o, by) {
  f  <- factor(o[[by, drop = TRUE]])
  mm <- Matrix::sparse.model.matrix(~ 0 + f)
  colnames(mm) <- levels(f)
  as.matrix(LayerData(o, assay = "RNA", layer = "counts") %*% mm)
}
# (Seurat's AggregateExpression(sub, group.by = "sample_id") does the same.)

pb <- pseudobulk(sub, "sample_id")
n_cells <- table(sub$sample_id)[colnames(pb)]
n_cells
keep_samples <- names(n_cells)[n_cells >= params$min_cells]
pb <- pb[, keep_samples]
coldata <- sample_info[colnames(pb), ]
coldata$n_cells <- as.vector(n_cells[colnames(pb)])

keep_genes <- filterByExpr(pb, group = coldata$group)
pb <- pb[keep_genes, ]
dim(pb)
```

```{r pb-pca, fig.width=11, fig.height=4.5}
dds <- DESeqDataSetFromMatrix(pb, colData = coldata, design = ~ sex + disease)
vsd <- vst(dds, blind = TRUE, nsub = min(1000, nrow(dds)))
plotPCA(vsd, intgroup = "disease") + ggtitle("PCA - disease") +
plotPCA(vsd, intgroup = "sex") + scale_colour_manual(values = sex_cols) + ggtitle("PCA - sex")
```

> **Exercise 3.3.** What separates the samples on PC1 and on PC2? Which
> genes load most on these PCs (`prcomp(t(assay(vsd)[top_var_genes, ]))`)?
> Are sex-chromosome genes among them?

```{r ex3.3-solution, eval=params$solutions, include=params$solutions}
top_var <- head(order(matrixStats::rowVars(assay(vsd)), decreasing = TRUE), 500)
pca <- prcomp(t(assay(vsd)[top_var, ]))
for (pc in c("PC1", "PC2")) {
  l <- sort(abs(pca$rotation[, pc]), decreasing = TRUE)[1:15]
  print(data.frame(gene = names(l), loading = round(l, 3), chr = chr_class(names(l)), row.names = NULL))
}
```

## 3.4 Model A: COVID-19 effect adjusted for sex, and sex effect adjusted for disease

```{r deseq-additive}
dds <- DESeq(dds, quiet = TRUE)
resultsNames(dds)

tidy_res <- function(res) {
  as.data.frame(res) %>% mutate(gene = rownames(res), chr = chr_class(rownames(res))) %>% arrange(pvalue)
}
res_disease <- tidy_res(results(dds, name = "disease_COVID19_vs_healthy"))
res_sex     <- tidy_res(results(dds, name = "sex_male_vs_female"))

n_sig <- function(r, alpha = 0.05) sum(r$padj < alpha, na.rm = TRUE)
c(disease_adj_for_sex = n_sig(res_disease), sex_adj_for_disease = n_sig(res_sex),
  cell_level_disease = sum(de_cell_disease$p_val_adj < 0.05))

head(res_disease, 10)
res_sex %>% filter(padj < 0.05) %>% count(chr)
```

```{r volcano, fig.width=12, fig.height=5}
volcano <- function(r, title, n_label = 12) {
  lab <- r %>% filter(padj < 0.05) %>% slice_min(pvalue, n = n_label)
  ggplot(r, aes(log2FoldChange, -log10(pvalue), colour = padj < 0.05 & !is.na(padj))) +
    geom_point(size = 0.8) +
    geom_text_repel(data = lab, aes(label = gene), size = 3, colour = "black", max.overlaps = 30) +
    scale_colour_manual(values = c(`FALSE` = "grey70", `TRUE` = "#CC3311"), name = "padj < 0.05") +
    ggtitle(title)
}
volcano(res_disease, paste(coi, "- COVID-19 vs healthy (adjusted for sex)")) +
volcano(res_sex,     paste(coi, "- male vs female (adjusted for disease)"))
```

> **Exercise 3.4.** Compare the number of DE genes between the per-cell
> test (3.1) and pseudobulk (3.4). Why is the per-cell number so much
> larger? Which do you trust more? (Squair *et al.*, 2021, *Nat Commun*;
> Zimmerman *et al.*, 2021, *Nat Commun*.)

```{r ex3.4-solution, eval=params$solutions, include=params$solutions}
cell_sig <- rownames(de_cell_disease)[de_cell_disease$p_val_adj < 0.05]
pb_sig   <- res_disease$gene[which(res_disease$padj < 0.05)]
c(cell_level = length(cell_sig), pseudobulk = length(pb_sig), shared = length(intersect(cell_sig, pb_sig)))
```

## 3.5 What happens if we forget about sex? A confounding experiment

Our design is balanced: both disease groups have 50% female samples. Many
real cohorts are not balanced. Imagine a study where all recruited
COVID-19 patients happened to be women and the healthy volunteer was a man.
We can mimic this by keeping only the **female COVID-19** samples and the
**male healthy** sample, and comparing COVID-19 vs healthy.

```{r confounding, fig.width=6, fig.height=5}
cd_conf <- coldata[(coldata$disease == "COVID19" & coldata$sex == "female") |
                   (coldata$disease == "healthy" & coldata$sex == "male"), ]
table(cd_conf$disease, cd_conf$sex)

dds_conf <- DESeq(DESeqDataSetFromMatrix(pb[, rownames(cd_conf)], cd_conf, design = ~ disease), quiet = TRUE)
res_conf <- tidy_res(results(dds_conf, name = "disease_COVID19_vs_healthy"))
head(res_conf[, c("gene", "chr", "log2FoldChange", "padj")], 15)
res_conf %>% filter(padj < 0.05) %>% count(chr) %>% mutate(pct = round(100 * n / sum(n), 1))
volcano(res_conf, "'COVID-19' vs healthy: female patients vs a male control")
```

> **Exercise 3.5.**
>
> 1. Which of the top "COVID-19 genes" are on the sex chromosomes? Compare
>    with `res_disease` (balanced design, adjusted for sex) from 3.4.
> 2. Can we fix this by adding `sex` to the design (`~ sex + disease`)?
>    Try it and read the error message.
> 3. Now keep the healthy female as well (COVID-19: 2 female; healthy: 1
>    female + 1 male) and fit `~ sex + disease`. What can the model estimate
>    now, and what is it still missing?
> 4. What does this mean for designing a study, and for reading published
>    ones?

```{r ex3.5-solution, eval=params$solutions, include=params$solutions}
# (2) sex and disease are perfectly confounded: the design matrix is not of full rank
tryCatch(DESeqDataSetFromMatrix(pb[, rownames(cd_conf)], cd_conf, design = ~ sex + disease),
         error = function(e) cat("Error:", conditionMessage(e), "\n"))

# (3) partial confounding: with a healthy woman the sex effect can be estimated (from the healthy donors only)
cd_part <- coldata[coldata$sex == "female" | coldata$disease == "healthy", ]
table(cd_part$disease, cd_part$sex)
res_part <- tidy_res(results(DESeq(DESeqDataSetFromMatrix(pb[, rownames(cd_part)], cd_part,
                                                          design = ~ sex + disease), quiet = TRUE),
                             name = "disease_COVID19_vs_healthy"))
head(res_part[, c("gene", "chr", "log2FoldChange", "padj")], 15)
# The model now assumes that the sex difference is the same in COVID-19 and healthy people (no interaction),
# and it can never check this, because there are no male patients.
# (4) Balance sex across groups at the DESIGN stage; always report the sex distribution per group; be
# suspicious of "disease genes" such as XIST, RPS4Y1 or DDX3Y.
```

## 3.6 Model B: sex-stratified effects and the interaction

We now give each sex × disease group its own mean (`~ 0 + group`). From these
means we build contrasts:

- **female:** female_COVID19 − female_healthy
- **male:** male_COVID19 − male_healthy
- **interaction:** (female effect) − (male effect)

```{r deseq-groups}
dds_g <- DESeqDataSetFromMatrix(pb, colData = coldata, design = ~ 0 + group)
dds_g <- DESeq(dds_g, quiet = TRUE)
resultsNames(dds_g)

contrast_vec <- function(plus, minus) {
  v <- setNames(rep(0, length(resultsNames(dds_g))), resultsNames(dds_g))
  v[paste0("group", plus)] <- 1
  v[paste0("group", minus)] <- -1
  v
}
res_F   <- tidy_res(results(dds_g, contrast = contrast_vec("female_COVID19", "female_healthy")))
res_M   <- tidy_res(results(dds_g, contrast = contrast_vec("male_COVID19",   "male_healthy")))
res_int <- tidy_res(results(dds_g, contrast = contrast_vec(c("female_COVID19", "male_healthy"),
                                                           c("female_healthy", "male_COVID19"))))
c(female = n_sig(res_F), male = n_sig(res_M), interaction = n_sig(res_int))

write.csv(res_F,   file.path(params$out_dir, paste0("DE_pseudobulk_", make.names(coi), "_female_COVIDvsHC.csv")), row.names = FALSE)
write.csv(res_M,   file.path(params$out_dir, paste0("DE_pseudobulk_", make.names(coi), "_male_COVIDvsHC.csv")),   row.names = FALSE)
write.csv(res_int, file.path(params$out_dir, paste0("DE_pseudobulk_", make.names(coi), "_interaction.csv")),      row.names = FALSE)
```

```{r fc-scatter, fig.width=7, fig.height=6}
cmp <- inner_join(select(res_F, gene, lfc_F = log2FoldChange, padj_F = padj),
                  select(res_M, gene, lfc_M = log2FoldChange, padj_M = padj), by = "gene") %>%
  left_join(select(res_int, gene, padj_int = padj), by = "gene") %>%
  mutate(across(starts_with("padj"), ~ replace_na(.x, 1)),
         category = case_when(padj_int < 0.05               ~ "interaction (sex-dependent)",
                              padj_F < 0.05 & padj_M < 0.05 ~ "both sexes",
                              padj_F < 0.05                 ~ "females only",
                              padj_M < 0.05                 ~ "males only",
                              TRUE                          ~ "not significant"))
table(cmp$category)

ggplot(cmp, aes(lfc_F, lfc_M, colour = category)) +
  geom_hline(yintercept = 0, colour = "grey80") + geom_vline(xintercept = 0, colour = "grey80") +
  geom_abline(linetype = "dashed", colour = "grey50") +
  geom_point(data = filter(cmp, category == "not significant"), size = 0.6) +
  geom_point(data = filter(cmp, category != "not significant"), size = 1.5) +
  geom_text_repel(data = cmp %>% filter(category != "not significant") %>% slice_min(pmin(padj_F, padj_M, padj_int), n = 15),
                  aes(label = gene), size = 3, show.legend = FALSE, max.overlaps = 30) +
  scale_colour_manual(values = c("both sexes" = "#CC3311", "females only" = "#E69F00", "males only" = "#009E73",
                                 "interaction (sex-dependent)" = "#332288", "not significant" = "grey80")) +
  labs(x = "log2FC COVID-19 vs healthy - females", y = "log2FC COVID-19 vs healthy - males",
       title = paste(coi, "- COVID-19 response by sex"))
```

> **Exercise 3.6 (key exercise).**
>
> 1. Pick a gene that is significant in "females only". Look up its
>    log2FC in males. Is it close to zero, or just not significant?
> 2. How many genes have a significant **interaction**? Compare with the
>    number of "females only" + "males only" genes. Which number would you
>    report as "sex-specific COVID-19 genes"?
> 3. Plot the expression of your gene per sample (e.g. `counts(dds_g,
>    normalized = TRUE)`), coloured by sex. What do you conclude?

```{r ex3.6-solution, eval=params$solutions, include=params$solutions, fig.width=5, fig.height=4}
f_only <- cmp %>% filter(category == "females only") %>% arrange(padj_F)
head(f_only, 10)
gene <- if (nrow(f_only) > 0) f_only$gene[1] else cmp$gene[which.min(cmp$padj_F)]
data.frame(norm_count = counts(dds_g, normalized = TRUE)[gene, ], coldata) %>%
  ggplot(aes(disease, norm_count, colour = sex)) +
  geom_point(size = 4, position = position_dodge(width = 0.4)) +
  scale_colour_manual(values = sex_cols) + ggtitle(gene)
# Most "females only" genes change in the same direction in males but miss the cut-off (dots near the
# diagonal in the scatter plot). Only the interaction test asks whether the effects DIFFER. Report that
# number - and keep in mind that with 2 residual df the interaction test has very little power.
```

```{r ex3.6-donor, eval=params$solutions && repeated_donors, include=params$solutions && repeated_donors}
# Advanced: repeated samples per donor -> limma-voom with the donor as a block
y <- calcNormFactors(DGEList(pb))
design_g <- model.matrix(~ 0 + group, data = coldata)
colnames(design_g) <- levels(coldata$group)
v   <- voom(y, design_g)
dc  <- duplicateCorrelation(v, design_g, block = coldata$donor)
v   <- voom(y, design_g, block = coldata$donor, correlation = dc$consensus.correlation)
fit <- lmFit(v, design_g, block = coldata$donor, correlation = dc$consensus.correlation)
cm  <- makeContrasts(female = female_COVID19 - female_healthy, male = male_COVID19 - male_healthy,
                     interaction = (female_COVID19 - female_healthy) - (male_COVID19 - male_healthy),
                     levels = design_g)
fit2 <- eBayes(contrasts.fit(fit, cm), robust = TRUE)
dc$consensus.correlation
summary(decideTests(fit2))
# Compare with DESeq2. (variancePartition::dream from Day 2 fits the donor as a proper random effect.)
```

> **Exercise 3.7 (donor-level pseudobulk).** Sum the pseudobulk counts of the
> two time points per donor (`pseudobulk(sub, "donor")`), so there is one
> profile per person. Fit `~ sex + disease` with DESeq2. How many DE genes
> do you get now? Then try `~ 0 + group`. What happens, and why?

```{r ex3.7-solution, eval=params$solutions && repeated_donors, include=params$solutions && repeated_donors}
pb_donor <- pseudobulk(sub, "donor")
cd_donor <- distinct(sample_info, donor, sex, disease, group)
rownames(cd_donor) <- cd_donor$donor
cd_donor <- cd_donor[colnames(pb_donor), ]
pb_donor <- pb_donor[filterByExpr(pb_donor, group = cd_donor$group), ]

dds_donor <- DESeq(DESeqDataSetFromMatrix(pb_donor, cd_donor, design = ~ sex + disease), quiet = TRUE)
c(disease = n_sig(tidy_res(results(dds_donor, name = "disease_COVID19_vs_healthy"))),
  sex     = n_sig(tidy_res(results(dds_donor, name = "sex_male_vs_female"))))

# one mean per group with one donor per group: 0 residual df, dispersion cannot be estimated
tryCatch(DESeq(DESeqDataSetFromMatrix(pb_donor, cd_donor, design = ~ 0 + group), quiet = TRUE),
         error = function(e) cat("Error:", conditionMessage(e), "\n"))
# The honest answer to "is the COVID-19 response different in women and men?" with these data is:
# we cannot test it. The sample-level results in 3.6 borrow their replication from the time points.
```

# 4. Enrichment analysis

Next we ask which biological processes are behind the DE genes, **separately
for each sex**. There are two approaches:

- **Over-representation analysis (ORA):** are GO terms over-represented
  among the significant genes? Needs a cut-off, and the background
  (*universe*) should be the genes we tested, not the whole genome.
- **Gene set enrichment analysis (GSEA):** uses all genes ranked by their
  test statistic, with no cut-off. Better when power is low, as here.

## 4.1 ORA: GO Biological Process, compared between the sexes

```{r ora}
universe <- rownames(dds_g)   # the genes we actually tested

gene_lists <- function(alpha) list(
  female_up   = res_F$gene[which(res_F$padj < alpha & res_F$log2FoldChange > 0)],
  female_down = res_F$gene[which(res_F$padj < alpha & res_F$log2FoldChange < 0)],
  male_up     = res_M$gene[which(res_M$padj < alpha & res_M$log2FoldChange > 0)],
  male_down   = res_M$gene[which(res_M$padj < alpha & res_M$log2FoldChange < 0)])

ora_lists <- Filter(function(g) length(g) >= 10, gene_lists(0.05))
if (length(ora_lists) == 0) {           # low power: fall back to a lenient, exploratory cut-off
  message("Fewer than 10 DE genes per list at padj < 0.05 - using padj < 0.2 (exploratory!)")
  ora_lists <- Filter(function(g) length(g) >= 10, gene_lists(0.2))
}
sapply(ora_lists, length)

ora <- NULL
if (length(ora_lists) > 0) {
  ora <- compareCluster(geneClusters = ora_lists, fun = "enrichGO",
                        OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP",
                        universe = universe, pAdjustMethod = "BH",
                        pvalueCutoff = 0.05, qvalueCutoff = 0.2, minGSSize = 10, maxGSSize = 500)
}
if (!is.null(ora) && nrow(as.data.frame(ora)) > 0) {
  write.csv(as.data.frame(ora), file.path(params$out_dir, paste0("GO_ORA_", make.names(coi), "_by_sex.csv")), row.names = FALSE)
  head(as.data.frame(ora))
} else message("No enriched GO terms")
```

```{r ora-plot, fig.width=9, fig.height=7, eval=!is.null(ora) && nrow(as.data.frame(ora)) > 0}
dotplot(ora, showCategory = 6) + ggtitle(paste(coi, "- GO BP (ORA), COVID-19 vs healthy by sex"))
```

> **Exercise 4.1.**
>
> 1. Which processes are enriched among the genes up-regulated in COVID-19?
>    Are they shared between the sexes?
> 2. Re-run `enrichGO` for `female_up` without the `universe` argument. How
>    do the results change, and why is the choice of universe important?
> 3. The female and male lists have different lengths. How can list size
>    alone create "sex-specific" terms?

```{r ex4.1-solution, eval=params$solutions && length(ora_lists) > 0, include=params$solutions && length(ora_lists) > 0}
ego_nouniv <- enrichGO(ora_lists[[1]], OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP",
                       pAdjustMethod = "BH", pvalueCutoff = 0.05)
ego_univ   <- enrichGO(ora_lists[[1]], OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP",
                       universe = universe, pAdjustMethod = "BH", pvalueCutoff = 0.05)
c(without_universe = nrow(as.data.frame(ego_nouniv)), with_universe = nrow(as.data.frame(ego_univ)))
# Without a universe, the background is every annotated gene in the genome, including genes that are not
# expressed in this cell type. Any list of immune-cell genes then looks "enriched" for immune processes.
```

## 4.2 GSEA: ranked genes, no cut-off

```{r gsea}
ranked <- list(female = sort(setNames(res_F$stat, res_F$gene), decreasing = TRUE),
               male   = sort(setNames(res_M$stat, res_M$gene), decreasing = TRUE))
ranked <- lapply(ranked, function(r) r[!is.na(r)])

gsea <- compareCluster(geneClusters = ranked, fun = "gseGO",
                       OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP",
                       minGSSize = 15, maxGSSize = 500, pvalueCutoff = 0.05, eps = 0, seed = TRUE)
gsea_df <- if (is.null(gsea)) data.frame() else as.data.frame(gsea)
nrow(gsea_df)
if (nrow(gsea_df) > 0) write.csv(gsea_df, file.path(params$out_dir, paste0("GO_GSEA_", make.names(coi), "_by_sex.csv")), row.names = FALSE)
```

```{r gsea-plot, fig.width=10, fig.height=8, eval=nrow(gsea_df) > 0}
dotplot(gsea, showCategory = 8, split = ".sign") + facet_grid(. ~ .sign) +
  ggtitle(paste(coi, "- GSEA GO BP, COVID-19 vs healthy by sex"))
```

```{r nes-compare, fig.width=7, fig.height=6, eval=nrow(gsea_df) > 0}
nes <- gsea_df %>% select(Cluster, ID, Description, NES, p.adjust) %>%
  pivot_wider(names_from = Cluster, values_from = c(NES, p.adjust))
ggplot(nes, aes(NES_female, NES_male)) +
  geom_hline(yintercept = 0, colour = "grey80") + geom_vline(xintercept = 0, colour = "grey80") +
  geom_abline(linetype = "dashed", colour = "grey50") + geom_point() +
  geom_text_repel(data = slice_max(nes, abs(coalesce(NES_female, 0)) + abs(coalesce(NES_male, 0)), n = 12),
                  aes(label = substr(Description, 1, 40)), size = 2.8) +
  labs(title = "GO terms significant in at least one sex", x = "NES females", y = "NES males")
```

> **Exercise 4.2.**
>
> 1. Do ORA and GSEA agree?
> 2. Is the interferon response (e.g. "response to virus", "type I
>    interferon signaling pathway") enriched in both sexes? Is it stronger in
>    one of them? (Compare with Takahashi *et al.*, 2020.)
> 3. A GO term missing from the male results (`NA` above) is not
>    necessarily "absent in males". How could you check this properly?
>    (Hint: `gseaplot2()`, or run GSEA on the interaction statistic
>    `res_int$stat`.)

```{r ex4.2-solution, eval=params$solutions, include=params$solutions, fig.width=9, fig.height=6}
ranked_int <- sort(setNames(res_int$stat, res_int$gene), decreasing = TRUE)
gsea_int <- gseGO(ranked_int[!is.na(ranked_int)], OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP",
                  minGSSize = 15, maxGSSize = 500, pvalueCutoff = 0.05, eps = 0, seed = TRUE)
head(as.data.frame(gsea_int)[, c("Description", "NES", "p.adjust")], 15)
# positive NES = more up-regulated by COVID-19 in females than in males
if (nrow(as.data.frame(gsea_int)) > 0) dotplot(gsea_int, showCategory = 10, split = ".sign") + facet_grid(. ~ .sign)
```

> **Exercise 4.3 (chromosomes).** Are the genes that differ by sex
> (`res_sex`, padj < 0.05) over-represented on the X chromosome? Use a
> Fisher test on a 2×2 table (DE yes/no × chrX yes/no), with all tested genes
> as background.

```{r ex4.3-solution, eval=params$solutions, include=params$solutions}
tab <- table(DE = res_sex$padj < 0.05 & !is.na(res_sex$padj), chrX = res_sex$chr == "chrX")
tab
fisher.test(tab)
```

# 5. Wrap-up

## What did we find?

```{r summary-table}
data.frame(
  analysis = c("DA: cell type chosen", "DA: selection rule",
               "DE per cell, COVID vs healthy", "DE pseudobulk, COVID vs healthy (adj. sex)",
               "DE pseudobulk, male vs female (adj. disease)",
               "DE pseudobulk, females: COVID vs healthy", "DE pseudobulk, males: COVID vs healthy",
               "DE pseudobulk, sex x disease interaction"),
  result = c(coi, rule, sum(de_cell_disease$p_val_adj < 0.05), n_sig(res_disease), n_sig(res_sex),
             n_sig(res_F), n_sig(res_M), n_sig(res_int)))
```

## A checklist for sex-sensitive single-cell studies (after SAGER)

- [ ] Report the number of **individuals** (not just cells or samples) per sex and group.
- [ ] Report how sex was determined (self-report, clinical record, genetics) and check it in the data (*XIST*/chrY).
- [ ] Use the correct terms: sex vs gender. If gender is relevant, collect it separately.
- [ ] Balance sex across the groups at the design stage, or at least model it.
- [ ] Treat individuals as the replicates (pseudobulk, mixed models), not cells.
- [ ] Test sex differences with an **interaction** term, not by comparing stratified p-values.
- [ ] Report sex-disaggregated results, even when not significant, so that future meta-analyses can use them.
- [ ] Discuss power: "no significant sex difference" ≠ "no sex difference".
- [ ] Watch out for sex-chromosome genes (and for loss of Y in older male donors).

## Final discussion

> Based on today's analysis, write two sentences you could put in the
> *Results* section of a paper about `r coi` in COVID-19, and one sentence
> for the *Limitations*.
>
> How many donors per group would you need for a real study? Look at the
> variability between the two healthy donors and between the time points of
> each patient, and discuss.

## Further reading

- Phipson B *et al.* (2022) propeller: testing for differences in cell type proportions in single cell data. *Bioinformatics*.
- Squair JW *et al.* (2021) Confronting false discoveries in single-cell differential expression. *Nat Commun*.
- Zimmerman KD, Espeland MA, Langefeld CD (2021) A practical solution to pseudoreplication bias in single-cell studies. *Nat Commun*.
- Crowell HL *et al.* (2020) muscat detects subpopulation-specific state transitions from multi-sample multi-condition single-cell transcriptomics data. *Nat Commun*.
- Gelman A, Stern H (2006) The difference between "significant" and "not significant" is not itself statistically significant. *Am Stat*.
- Heidari S *et al.* (2016) Sex and Gender Equity in Research: rationale for the SAGER guidelines. *Res Integr Peer Rev*.
- Klein SL, Flanagan KL (2016) Sex differences in immune responses. *Nat Rev Immunol*.
- Takahashi T *et al.* (2020) Sex differences in immune responses that underlie COVID-19 disease outcomes. *Nature*.
- Tukiainen T *et al.* (2017) Landscape of X chromosome inactivation across human tissues. *Nature*.

```{r session-info}
sessionInfo()
```
