---
title: "Single-cell transcriptomics with a sex & gender lens"
subtitle: "Part 1 - from count matrices to annotated cells (6 PBMC samples)"
author: "Joana P. Bernardes & Neha Mishra"
date: "MOSAIC-ID workshop 2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_download: true
params:
  input_dir: "input"               # folder that contains one sub-folder per sample (Cell Ranger output)
  sample_sheet: "sample_sheet.csv" # one row per sample: sample_id, folder, donor, disease, sex, timepoint
  ref_dir: "~/References"          # PBMC reference, housekeeping genes, marker list, SingleR references
  out_dir: "output"
  solutions: false                 # TRUE renders the instructor version (with solutions)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(33)
```

# Introduction

In this practical you will perform the first steps of a single-cell RNA-seq
(scRNA-seq) analysis. Throughout, we will keep asking one question:
**does sex matter here, and how would we know?**

Part 1 (this notebook):

- Quality control (QC) of the cells
- Checking the recorded sex of each sample in the data
- Normalisation, integration of samples, dimensionality reduction
- Clustering and cell-type annotation

Part 2 (`D3_singleCell_Sex_downstream_2026.Rmd`) covers the downstream analyses
with a sex-sensitive design:

- Differential abundance (cell proportions)
- Differential expression, first per cell and then as pseudobulk
- Enrichment analysis

## Sex and gender: what can we measure?

- **Sex** refers to biological attributes: sex chromosomes, gene expression,
  hormones and anatomy. Sex is not always binary (e.g. XXY, XO, intersex
  variations).
- **Gender** refers to socially constructed roles, behaviours and identities.
  It shapes exposures, health behaviour and access to care, and through those
  it can affect immunity.

A transcriptome can tell us something about **sex chromosomes**
(e.g. *XIST*, chrY genes). It can tell us **nothing about gender**. In clinical
metadata the column is often called "gender" when it actually holds sex (as it
did in the 2022 version of this course!). The SAGER guidelines
(Heidari *et al.*, 2016, *Res Integr Peer Rev*) ask us to use these terms
correctly, to report how sex was determined, and to report results
disaggregated by sex.

## Why a sex lens for COVID-19?

- Men have a higher risk of severe COVID-19 and of death.
- Female and male patients mount different immune responses. Takahashi *et al.*
  (2020, *Nature*) found that male patients had higher innate cytokines and
  more non-classical monocytes, while female patients had stronger T-cell
  activation.
- Several immune genes lie on the X chromosome (e.g. *TLR7*, *CD40LG*,
  *CXCR3*). Some of them can escape X inactivation, so females can express
  them from both X copies.
- For reviews see Klein & Flanagan (2016, *Nat Rev Immunol*) and Scully *et al.*
  (2020, *Nat Rev Immunol*).

## Our data

PBMCs (peripheral blood mononuclear cells) from Bernardes *et al.* (2020,
*Immunity*), 10x Genomics 3' scRNA-seq. The data are available upon request.

| | female | male |
|---|---|---|
| COVID-19 | 1 donor, 2 time points (2 samples) | 1 donor, 2 time points (2 samples) |
| healthy | 1 donor (1 sample) | 1 donor (1 sample) |

That is 6 samples, but only **4 people**: one per sex × disease group.

> **Exercise 0 (discussion, 5 min).** Look at `sample_sheet.csv`.
> How many *samples* do we have per group? How many *individuals (donors)*?
> Which of the two counts is our number of biological replicates? Write down
> what you expect you *can* and *cannot* conclude about sex differences with
> this design. We will come back to your answer in Part 2.

```{asis, echo=params$solutions}
**Discussion notes (instructor).** 6 samples, 4 donors, 1 donor per group:
the biological replicate is the donor, so n = 1 per group. The two time
points of a COVID-19 patient are *repeated measures* of the same person.
Every comparison between groups is also a comparison between individuals:
"female vs male" in the healthy group is one woman vs one man. We can
**learn and practise** sex-sensitive methods on these data, but we cannot
generalise sex differences to a population. That needs several donors per
group (see Part 2).
```

In this tutorial we use the R package **Seurat** (v5). Popular alternatives
include scanpy (Python), scran/scater (Bioconductor) and the book
[Orchestrating Single-Cell Analysis with Bioconductor (OSCA)](https://bioconductor.org/books/release/OSCA/).
A list of single-cell tools is maintained
[here](https://github.com/seandavi/awesome-single-cell).

---

# 0. Setup

```{r libraries}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(RColorBrewer)
library(org.Hs.eg.db)

# Bioconductor packages mask some dplyr verbs (e.g. AnnotationDbi::select, matrixStats::count)
select <- dplyr::select; count <- dplyr::count; filter <- dplyr::filter

dir.create(params$out_dir, showWarnings = FALSE, recursive = TRUE)
ref_dir <- path.expand(params$ref_dir)
```

# 1. Load the data

## 1.1 The sample sheet

We describe every sample **once** in a sample sheet instead of typing the
metadata by hand for each object. This is less work and avoids copy-paste
errors, such as assigning the wrong sex to a sample.

```{r sample-sheet}
# fileEncoding handles CSVs saved from Excel; strip.white removes stray spaces
samples <- read.csv(params$sample_sheet, stringsAsFactors = FALSE,
                    fileEncoding = "UTF-8-BOM", strip.white = TRUE)
samples

required <- c("sample_id", "folder", "donor", "disease", "sex", "timepoint")
missing_cols <- setdiff(required, colnames(samples))
if (length(missing_cols) > 0) stop("Sample sheet is missing column(s): ", paste(missing_cols, collapse = ", "),
                                   ". Found: ", paste(colnames(samples), collapse = ", "))
stopifnot(
  "Please fill in the FILL_ME fields of the sample sheet" = !any(samples == "FILL_ME", na.rm = TRUE),
  "sex must be 'female' or 'male'"         = all(samples$sex %in% c("female", "male")),
  "disease must be 'healthy' or 'COVID19'" = all(samples$disease %in% c("healthy", "COVID19")),
  "sample_id must be unique"               = !any(duplicated(samples$sample_id))
)

table(samples$disease, samples$sex)                  # samples per group
table(unique(samples[, c("donor", "disease", "sex")])[, c("disease", "sex")]) # donors per group
```

## 1.2 Read the 10x count matrices

Cell Ranger writes two matrices per sample:

- `raw_feature_bc_matrix`: **all** barcodes, including empty droplets
- `filtered_feature_bc_matrix`: only barcodes that Cell Ranger called as cells

In this course we start from the **raw** matrix of every sample. Most of its
barcodes are empty droplets that contain only ambient RNA. `min.features = 200`
in `CreateSeuratObject()` removes barcodes with fewer than 200 detected genes,
which drops most of them. That is a crude filter; `DropletUtils::emptyDrops()`
is the proper tool.

```{r read10x}
read_sample <- function(folder) {
  # `folder` can be a sub-folder of params$input_dir or a full path to the sample folder
  folder <- sub("/+$", "", path.expand(folder))
  base <- if (dir.exists(folder)) folder else file.path(params$input_dir, folder)
  mdir <- if (basename(base) == "raw_feature_bc_matrix") base else file.path(base, "raw_feature_bc_matrix")
  if (!dir.exists(mdir)) stop("No raw_feature_bc_matrix folder found for ", folder, " (looked in ", mdir, ")")
  message("Reading ", mdir)
  counts <- Read10X(data.dir = mdir)
  if (is.list(counts)) counts <- counts[["Gene Expression"]]  # CITE-seq / multiome output
  counts
}

seurat_list <- lapply(seq_len(nrow(samples)), function(i) {
  counts <- read_sample(samples$folder[i])
  obj <- CreateSeuratObject(counts = counts,
                            project = samples$sample_id[i],
                            min.cells = 3, min.features = 200)
  obj@misc$n_raw_barcodes <- ncol(counts)
  # add the sample-level metadata to every cell
  for (col in c("sample_id", "donor", "disease", "sex", "timepoint")) obj[[col]] <- samples[i, col]
  obj
})
names(seurat_list) <- samples$sample_id
names(seurat_list)   # the objects are now called by their sample_id, e.g. seurat_list$HC_M

data.frame(raw_barcodes       = sapply(seurat_list, function(o) o@misc$n_raw_barcodes),
           after_min_features = sapply(seurat_list, ncol))  # barcodes kept per sample
```

> **Question.** How many barcodes does the raw matrix of each sample contain,
> and how many are left after `min.features = 200`? Compare with the number of cells Cell Ranger reports
> in each sample's `web_summary.html`.

> **Exercise 1.1.** Pick one object, e.g. `seurat_list$HC_M`.
> What does `dim()` return? Which number is cells and which is genes?
> Where are the counts stored (`Layers()`, `LayerData()`)? Where is the
> cell metadata stored (`@meta.data`)?

```{r ex1.1-solution, eval=params$solutions, include=params$solutions}
dim(seurat_list$HC_M)                  # genes x cells
Layers(seurat_list$HC_M)               # "counts"
LayerData(seurat_list$HC_M, layer = "counts")[1:5, 1:5]
head(seurat_list$HC_M@meta.data)
```

## 1.3 Merge the samples into one object

`merge()` keeps the samples as separate *layers*, which we need later for
integration. We prefix the cell barcodes with the sample ID so that barcodes
are unique across samples.

```{r merge}
obj <- merge(x = seurat_list[[1]], y = seurat_list[-1],
             add.cell.ids = names(seurat_list), project = "D3_sex")
rm(seurat_list); invisible(gc())
obj
Layers(obj)
```

# 2. Quality control

## 2.1 QC metrics

- `nFeature_RNA`: number of genes detected per cell
- `nCount_RNA`: number of UMIs per cell
- `percent.mt`: % of UMIs from mitochondrial genes (high in stressed or dying cells)
- `percent.ribo`: % of UMIs from ribosomal protein genes
- `percent.hk`: % of UMIs from housekeeping genes (low in poor-quality cells)

```{r qc-metrics}
obj[["percent.mt"]]   <- PercentageFeatureSet(obj, pattern = "^MT-")
obj[["percent.ribo"]] <- PercentageFeatureSet(obj, pattern = "^RP[SL]")

hk_file <- file.path(ref_dir, "Housekeeping_Hs.csv")
if (file.exists(hk_file)) {
  hkgenes <- toupper(read.csv(hk_file, header = TRUE)$V1)
  hkgenes.found <- rownames(obj)[toupper(rownames(obj)) %in% hkgenes]
  obj[["percent.hk"]] <- PercentageFeatureSet(obj, features = hkgenes.found)
} else {
  message("Housekeeping list not found - percent.hk will not be used")
}
qc_features <- intersect(c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.ribo", "percent.hk"),
                         colnames(obj@meta.data))
```

> **Question.** Where were `percent.mt` and `percent.hk` saved in the object?

## 2.2 Visualise the unfiltered data

```{r qc-vln-before, fig.width=14, fig.height=4}
VlnPlot(obj, features = qc_features, group.by = "sample_id",
        ncol = length(qc_features), pt.size = 0)
```

```{r qc-scatter, fig.width=10, fig.height=4}
FeatureScatter(obj, "nCount_RNA", "nFeature_RNA", group.by = "sample_id", raster = TRUE) +
FeatureScatter(obj, "nCount_RNA", "percent.mt",   group.by = "sample_id", raster = TRUE)
```

> **Exercise 2.1 (sex lens on QC).** Plot the QC metrics grouped by `sex`
> and by `disease` instead of `sample_id`. Do female and male samples
> differ? If they did, would it be technical (a batch processed on another
> day) or biological? Why does it matter that we use the **same** thresholds
> for all groups?

```{r ex2.1-solution, eval=params$solutions, include=params$solutions, fig.width=14, fig.height=4}
VlnPlot(obj, features = qc_features, group.by = "sex",     ncol = length(qc_features), pt.size = 0)
VlnPlot(obj, features = qc_features, group.by = "disease", ncol = length(qc_features), pt.size = 0)
obj@meta.data %>% group_by(sample_id, sex, disease) %>%
  summarise(across(all_of(qc_features), median), n_cells = n(), .groups = "drop")
```

## 2.3 Filter: keep high-quality cells

```{r qc-filter}
max_features <- 5000   # very high numbers of genes may indicate doublets
min_features <- 200
max_mt       <- 25     # PBMCs from COVID-19 patients can have more mt reads - check your plots!
min_hk       <- 10

cells_before <- table(obj$sample_id)
keep <- obj$nFeature_RNA > min_features & obj$nFeature_RNA < max_features & obj$percent.mt < max_mt
if ("percent.hk" %in% qc_features) keep <- keep & obj$percent.hk > min_hk
obj <- subset(obj, cells = colnames(obj)[keep])

data.frame(before = as.vector(cells_before),
           after  = as.vector(table(obj$sample_id)[names(cells_before)]),
           row.names = names(cells_before)) %>%
  mutate(pct_kept = round(100 * after / before, 1))
```

```{r qc-vln-after, fig.width=14, fig.height=4}
VlnPlot(obj, features = qc_features, group.by = "sample_id",
        ncol = length(qc_features), pt.size = 0)
```

> **Exercise 2.2.** Fixed thresholds are a choice. An alternative is
> *per-sample, data-driven* thresholds, e.g. removing cells more than 3
> median absolute deviations (MADs) from the sample median. Compute the
> 3-MAD upper limit of `percent.mt` for each sample. Would it change which
> cells are removed? Does it remove more cells from one sex or one disease
> group?

```{r ex2.2-solution, eval=params$solutions, include=params$solutions}
obj@meta.data %>%
  group_by(sample_id, sex, disease) %>%
  summarise(median_mt = median(percent.mt),
            mad3_upper = median(percent.mt) + 3 * mad(percent.mt),
            pct_above  = round(100 * mean(percent.mt > median(percent.mt) + 3 * mad(percent.mt)), 1),
            .groups = "drop")
```

# 3. Check the recorded sex in the data

Sample swaps and wrong metadata are common. Our samples come from different
individuals, so we can check the recorded sex with sex-chromosome genes:

- *XIST* (X-inactive specific transcript) is expressed from the inactive X, so
  only in cells with two or more X chromosomes.
- chrY genes (*RPS4Y1*, *DDX3Y*, *UTY*, *KDM5D*, *EIF1AY*, ...) are expressed
  only when a Y chromosome is present.

```{r normalize}
obj <- NormalizeData(obj)
```

```{r sex-check, fig.width=11, fig.height=4}
y_genes <- intersect(c("RPS4Y1", "DDX3Y", "UTY", "KDM5D", "EIF1AY", "USP9Y", "ZFY", "TXLNGY", "NLGN4Y"),
                     rownames(obj))
obj <- JoinLayers(obj)            # one matrix for all cells (we split again for integration)
expr <- LayerData(obj, layer = "data")
obj$XIST_expr  <- expr["XIST", ]
obj$chrY_score <- Matrix::colMeans(expr[y_genes, , drop = FALSE])

sex_check <- obj@meta.data %>%
  group_by(sample_id, donor, recorded_sex = sex) %>%
  summarise(XIST = mean(XIST_expr), chrY = mean(chrY_score),
            pct_XIST_pos = round(100 * mean(XIST_expr > 0), 1),
            pct_chrY_pos = round(100 * mean(chrY_score > 0), 1), .groups = "drop") %>%
  mutate(inferred_sex = case_when(XIST > chrY ~ "female", chrY > XIST ~ "male", TRUE ~ "unclear"),
         match = inferred_sex == recorded_sex)
sex_check

ggplot(sex_check, aes(XIST, chrY, colour = recorded_sex, label = sample_id)) +
  geom_point(size = 4) + geom_text(vjust = -1, size = 3) +
  scale_colour_manual(values = c(female = "#E69F00", male = "#009E73")) +
  labs(x = "mean XIST (log-normalised)", y = "mean chrY genes", title = "Sample-level sex check") +
  theme_bw() |
VlnPlot(obj, features = c("XIST", "RPS4Y1"), group.by = "sample_id", pt.size = 0, ncol = 2)

if (!all(sex_check$match)) warning("Recorded and inferred sex disagree for: ",
                                   paste(sex_check$sample_id[!sex_check$match], collapse = ", "))
```

> **Exercise 3.1 (discussion).**
>
> 1. Do the recorded and the inferred sex agree for all samples?
> 2. Some cells from male samples show a little *XIST*, and some cells
>    from female samples show chrY reads. Where could that come from?
>    (Think of ambient RNA, doublets, and pooled or multiplexed samples.)
> 3. Blood cells of older men often lose the Y chromosome (*mosaic loss of
>    chromosome Y*, LOY). How would LOY show up in this plot?
> 4. What would you conclude about a sample with high *XIST* **and** high
>    chrY expression (e.g. XXY, Klinefelter syndrome)? About a trans woman
>    on hormone therapy?
> 5. Can we infer **gender** from these data?

```{asis, echo=params$solutions}
**Discussion notes (instructor).** (2) Ambient RNA in the droplets and
doublets give low-level off-target signal; pooled samples from mixed sexes
would give many cells of each kind. (3) With LOY, the fraction of chrY+
cells drops in male samples while *XIST* stays off. (4) Sex-chromosome
genes tell us karyotype-related expression only. Hormone therapy does not
change the karyotype; hormone-responsive genes may change. (5) No: gender is
social and cannot be read from a transcriptome. We must ask people, and
record the answer with their consent.
```

# 4. Normalisation, variable genes and integration

## 4.1 Variable genes: keep sex-chromosome genes out

*XIST* and the chrY genes differ strongly between female and male cells. If
they are among the highly variable genes (HVGs), some principal components
can split cells **by sex** instead of by cell type. Here we want clusters to
be cell types, so we remove *XIST*/*TSIX* and chrY genes from the HVGs. They
stay in the data for the downstream analyses.

```{r gene-chr}
# chromosome of each gene (used again in Part 2)
valid <- intersect(rownames(obj), keys(org.Hs.eg.db, keytype = "SYMBOL"))
gene_chr <- AnnotationDbi::mapIds(org.Hs.eg.db, keys = valid, keytype = "SYMBOL", column = "CHR",
                                  multiVals = function(x) paste(sort(unique(x)), collapse = ";"))
gene_chr <- setNames(ifelse(is.na(gene_chr[rownames(obj)]), "unknown", gene_chr[rownames(obj)]), rownames(obj))
table(gene_chr)[c("X", "Y", "X;Y")]   # "X;Y" = pseudo-autosomal genes
sex_chr_genes <- names(gene_chr)[gene_chr %in% c("Y", "X;Y") | names(gene_chr) %in% c("XIST", "TSIX")]
```

```{r split-hvg}
obj[["RNA"]] <- split(obj[["RNA"]], f = obj$sample_id)  # one layer per sample for integration
obj <- FindVariableFeatures(obj, selection.method = "vst", nfeatures = 2000)

intersect(VariableFeatures(obj), sex_chr_genes)          # which sex genes were highly variable?
VariableFeatures(obj) <- setdiff(VariableFeatures(obj), sex_chr_genes)
length(VariableFeatures(obj))
```

> **Exercise 4.1.** Which sex-chromosome genes were among the HVGs?
> Skip the `setdiff()` step, run the PCA below and colour the PCs by `sex`
> (`DimPlot(obj, reduction = "pca", group.by = "sex", dims = c(i, j))` or
> `DimHeatmap()`). Does any PC separate the sexes? Is removing these genes
> always a good idea? When would you keep them?

## 4.2 Scale, PCA and the unintegrated UMAP

> **Question.** What does `ScaleData()` actually do (`?ScaleData`)? Why do we
> only scale the variable genes?

```{r pca, fig.width=6, fig.height=4}
obj <- ScaleData(obj)
obj <- RunPCA(obj, npcs = 50, verbose = FALSE)
ElbowPlot(obj, ndims = 50)
```

```{r umap-unintegrated, fig.width=12, fig.height=5}
n_pcs <- 30
obj <- RunUMAP(obj, dims = 1:n_pcs, reduction = "pca", reduction.name = "umap.unintegrated")
DimPlot(obj, reduction = "umap.unintegrated", group.by = "sample_id") +
DimPlot(obj, reduction = "umap.unintegrated", group.by = "sex")
```

## 4.3 Integration of the 6 samples

Every sample was processed separately, so part of the variation between
samples is technical (a batch effect). Integration aligns the shared cell
types across samples. We use reciprocal PCA (RPCA), which is fast. CCA (used in
the 2022 course) and Harmony are alternatives.

**Caution:** each sample is its own batch *and* its own donor, sex and disease
group. Integration may therefore also remove some real biology. We use
the integrated space **only** for clustering and visualisation. All
differential analyses in Part 2 use the uncorrected counts.

```{r integrate}
obj <- IntegrateLayers(obj, method = RPCAIntegration, orig.reduction = "pca",
                       new.reduction = "integrated.rpca", verbose = FALSE)
obj <- JoinLayers(obj)   # re-join the layers after integration
obj <- RunUMAP(obj, reduction = "integrated.rpca", dims = 1:n_pcs, reduction.name = "umap")
```

```{r umap-integrated, fig.width=15, fig.height=4.5}
DimPlot(obj, reduction = "umap", group.by = "sample_id") +
DimPlot(obj, reduction = "umap", group.by = "sex", cols = c(female = "#E69F00", male = "#009E73")) +
DimPlot(obj, reduction = "umap", group.by = "disease")
```

> **Exercise 4.2.** Compare the unintegrated and the integrated UMAP. Which
> samples looked most different before integration? After integration, do
> female and male cells mix within every region of the UMAP? Where do they not
> mix, and what could that mean?

```{r ex4.2-solution, eval=params$solutions, include=params$solutions, fig.width=10, fig.height=5}
DimPlot(obj, reduction = "umap", split.by = "sex", group.by = "disease")
```

## 4.4 Clustering

```{r clusters, fig.width=15, fig.height=5}
obj <- FindNeighbors(obj, reduction = "integrated.rpca", dims = 1:n_pcs)
for (res in c(0.1, 0.2, 0.3)) obj <- FindClusters(obj, resolution = res, verbose = FALSE)
DimPlot(obj, group.by = "RNA_snn_res.0.1", label = TRUE) + NoLegend() + ggtitle("res 0.1") |
DimPlot(obj, group.by = "RNA_snn_res.0.2", label = TRUE) + NoLegend() + ggtitle("res 0.2") |
DimPlot(obj, group.by = "RNA_snn_res.0.3", label = TRUE) + NoLegend() + ggtitle("res 0.3")
```

> **Exercise 4.3.** What happens to the clusters when you change the
> resolution? Which resolution would you choose, and why? Try 0.5 and 1.

**IMPORTANT: choose a cluster resolution.**

```{r choose-res, fig.width=6, fig.height=5}
obj$seurat_clusters <- obj$RNA_snn_res.0.2
Idents(obj) <- "seurat_clusters"
DimPlot(obj, reduction = "umap", label = TRUE)
```

## 4.5 Sample and sex distribution per cluster

```{r cluster-composition, fig.width=12, fig.height=4}
ggplot(obj@meta.data, aes(x = seurat_clusters, fill = sample_id)) +
  geom_bar(position = "fill") + ylab("fraction of cells") + theme_bw() +
ggplot(obj@meta.data, aes(x = seurat_clusters, fill = sex)) +
  geom_bar(position = "fill") + ylab("fraction of cells") + theme_bw() +
  scale_fill_manual(values = c(female = "#E69F00", male = "#009E73"))
```

```{r cluster-qc, fig.width=15, fig.height=4}
VlnPlot(obj, features = qc_features, ncol = length(qc_features), pt.size = 0)
```

> **Question.** Is any cluster made up mostly of one sample or one sex? Is
> any cluster defined by poor QC (high `percent.mt`, low `nFeature_RNA`)?

# 5. Cell-type annotation

## 5.1 Reference mapping (Azimuth PBMC reference, Hao *et al.* 2021)

```{r ref-mapping, fig.width=14, fig.height=5}
ref_file <- file.path(ref_dir, "PBMCs_SeuratV4.rda")
has_reference <- file.exists(ref_file)

if (has_reference) {
  load(ref_file)                                   # loads the object `reference`
  reference <- UpdateSeuratObject(reference)
  query <- SCTransform(obj, verbose = FALSE)
  anchors <- FindTransferAnchors(reference = reference, query = query,
                                 normalization.method = "SCT",
                                 reference.reduction = "spca",
                                 recompute.residuals = FALSE, dims = 1:50)
  query <- MapQuery(anchorset = anchors, query = query, reference = reference,
                    refdata = list(celltype.l1 = "celltype.l1", celltype.l2 = "celltype.l2"),
                    reference.reduction = "spca", reduction.model = "wnn.umap")
  obj$celltype.l1 <- query$predicted.celltype.l1
  obj$celltype.l2 <- query$predicted.celltype.l2
  obj$celltype.l2.score <- query$predicted.celltype.l2.score
  rm(query, anchors, reference); invisible(gc())

  print(DimPlot(obj, group.by = "celltype.l1", label = TRUE, repel = TRUE) +
        DimPlot(obj, group.by = "celltype.l2", label = FALSE))
} else {
  message("Reference not found at ", ref_file, " - skipping reference mapping.")
}
```

Alternative: `Azimuth::RunAzimuth(obj, reference = "pbmcref")` downloads the
same reference and runs these steps in one call.

## 5.2 SingleR (optional)

```{r singler, eval = requireNamespace("SingleR", quietly = TRUE) && file.exists(file.path(path.expand(params$ref_dir), "Reference_dmap.RDS")), fig.width=12, fig.height=5}
library(SingleR)
dmap.se <- readRDS(file.path(ref_dir, "Reference_dmap.RDS"))
hpca.se <- readRDS(file.path(ref_dir, "Reference_hpca.RDS"))

input <- LayerData(obj, assay = "RNA", layer = "data")  # keep it sparse: as.matrix() needs >10 GB here
obj$hpca.labels <- SingleR(test = input, ref = hpca.se, labels = hpca.se$label.main)$labels
obj$dmap.labels <- SingleR(test = input, ref = dmap.se, labels = dmap.se$label.main)$labels
rm(input)

DimPlot(obj, group.by = "hpca.labels", label = TRUE, repel = TRUE) +
DimPlot(obj, group.by = "dmap.labels")
```

## 5.3 Marker genes

```{r markers, fig.width=9, fig.height=10}
canonical_markers <- list(
  "CD4+ T cells"       = c("CD3E", "CD4", "IL7R"),
  "CD8+ T cells"       = c("CD3E", "CD8A", "CD8B"),
  "other T cells"      = c("TRDC", "TRGC1", "SLC4A10"),
  "NK cells"           = c("NKG7", "GNLY", "KLRD1"),
  "B cells"            = c("MS4A1", "CD79A", "CD19"),
  "Plasmablasts"       = c("MZB1", "JCHAIN", "XBP1"),
  "CD14 Monocytes"     = c("CD14", "LYZ", "S100A8"),
  "CD16 Monocytes"     = c("FCGR3A", "MS4A7", "CDKN1C"),
  "DC"                 = c("FCER1A", "CD1C", "CLEC10A", "LILRA4"),
  "Megakaryocytes"     = c("PPBP", "PF4", "GP9"),
  "Prolif Lymphocytes" = c("MKI67", "TOP2A"),
  "HSPC"               = c("CD34")
)
canonical_markers <- Filter(length, lapply(canonical_markers, intersect, rownames(obj)))

marker_file <- file.path(ref_dir, "PBMCs_marker.csv")
marker_genes <- if (file.exists(marker_file)) {
  read.csv(marker_file, header = TRUE, sep = ";")$GeneMarker
} else unlist(canonical_markers)
marker_genes <- unique(intersect(marker_genes, rownames(obj)))

color <- c("#B2182B", "#D6604D", "#F4A582", "#FDDBC7", "#F7F7F7", "#D1E5F0", "#92C5DE", "#4393C3", "#2166AC")
DotPlot(obj, features = marker_genes, group.by = "seurat_clusters", dot.scale = 6) +
  scale_color_gradientn(colours = rev(color)) + coord_flip() + scale_y_discrete(position = "right") +
  theme(axis.text.x = element_text(size = 12))
```

## 5.4 Combine the evidence

```{r confusion, fig.width=12, fig.height=6, eval=has_reference}
cells_cluster <- table(cluster = obj$seurat_clusters, label = obj$celltype.l2)
scaled_cM <- round(100 * prop.table(cells_cluster, margin = 1), 1)
scaled_cM <- scaled_cM[, colSums(scaled_cM) > 0, drop = FALSE]

tmp <- as.data.frame.matrix(scaled_cM[, colSums(scaled_cM) >= 5, drop = FALSE])
tmp$others <- rowSums(scaled_cM[, colSums(scaled_cM) < 5, drop = FALSE])

pheatmap::pheatmap(as.matrix(tmp), border_color = "black", display_numbers = TRUE,
                   number_color = "black", color = c("white", brewer.pal(9, "YlOrRd")),
                   cluster_cols = FALSE, cluster_rows = FALSE,
                   main = "% of cells in each cluster (rows) per reference label (columns)")
```

> **Exercise 5.1.** Use the DotPlot and the heatmap to name each cluster
> with one of: "CD4+ T cells", "CD8+ T cells", "other T cells", "NK cells",
> "Prolif Lymphocytes", "B cells", "Plasmablasts", "CD14 Monocytes",
> "CD16 Monocytes", "DC", "Megakaryocytes", "HSPC", "Doublets".
> Fill in the vector below (one name per cluster, in cluster order), then
> compare your annotation with the automatic one in 5.5.

```{r ex5.1, eval=FALSE}
levels(obj$seurat_clusters)
my.cluster.ids <- c("...", "...", "...")   # one entry per cluster
names(my.cluster.ids) <- levels(obj$seurat_clusters)
obj$my_CellTypes <- unname(my.cluster.ids[as.character(obj$seurat_clusters)])
table(obj$my_CellTypes, obj$CellTypes)       # run after 5.5
```

## 5.5 Final annotation

So that everybody continues with the same labels, we annotate each cluster
automatically:

- With the reference: the most frequent `celltype.l2` label in the cluster,
  collapsed to the broad cell types above.
- Without it: the cell type whose canonical markers score highest in the
  cluster.

**Instructors: always check this against the DotPlot!** Correct a cluster by
hand if needed (see the commented line).

```{r final-annotation}
l2_to_broad <- c(
  "CD4 Naive" = "CD4+ T cells", "CD4 TCM" = "CD4+ T cells", "CD4 TEM" = "CD4+ T cells",
  "CD4 CTL" = "CD4+ T cells", "Treg" = "CD4+ T cells",
  "CD8 Naive" = "CD8+ T cells", "CD8 TCM" = "CD8+ T cells", "CD8 TEM" = "CD8+ T cells",
  "gdT" = "other T cells", "MAIT" = "other T cells", "dnT" = "other T cells",
  "CD4 Proliferating" = "Prolif Lymphocytes", "CD8 Proliferating" = "Prolif Lymphocytes",
  "NK Proliferating" = "Prolif Lymphocytes",
  "NK" = "NK cells", "NK_CD56bright" = "NK cells", "ILC" = "NK cells",
  "B naive" = "B cells", "B intermediate" = "B cells", "B memory" = "B cells",
  "Plasmablast" = "Plasmablasts",
  "CD14 Mono" = "CD14 Monocytes", "CD16 Mono" = "CD16 Monocytes",
  "cDC1" = "DC", "cDC2" = "DC", "pDC" = "DC", "ASDC" = "DC",
  "Platelet" = "Megakaryocytes", "Eryth" = "Erythrocytes", "HSPC" = "HSPC", "Doublet" = "Doublets")

if (has_reference) {
  obj$broad_ref <- unname(l2_to_broad[obj$celltype.l2])
  cluster_annotation <- obj@meta.data %>% count(seurat_clusters, broad_ref) %>%
    group_by(seurat_clusters) %>% mutate(purity = round(100 * n / sum(n), 1)) %>%
    slice_max(n, n = 1, with_ties = FALSE) %>% select(seurat_clusters, CellType = broad_ref, purity)
} else {
  obj <- AddModuleScore(obj, features = canonical_markers, name = "marker_score")
  score_cols <- paste0("marker_score", seq_along(canonical_markers))
  mean_scores <- obj@meta.data %>% group_by(seurat_clusters) %>%
    summarise(across(all_of(score_cols), mean))
  cluster_annotation <- data.frame(
    seurat_clusters = mean_scores$seurat_clusters,
    CellType = names(canonical_markers)[max.col(as.matrix(mean_scores[, score_cols]))])
}
cluster_annotation

new.cluster.ids <- setNames(cluster_annotation$CellType, cluster_annotation$seurat_clusters)
# new.cluster.ids["5"] <- "Doublets"      # manual correction example
obj$CellTypes <- unname(new.cluster.ids[as.character(obj$seurat_clusters)])
Idents(obj) <- "CellTypes"
table(obj$CellTypes, obj$sample_id)
```

```{r remove-doublets, fig.width=7, fig.height=5}
obj <- obj[, !obj$CellTypes %in% c("Doublets", "Erythrocytes")]

celltype_colors <- c("CD4+ T cells" = "#EA811F", "CD8+ T cells" = "maroon", "other T cells" = "peru",
                     "NK cells" = "#6A3D8A", "B cells" = "goldenrod1", "Plasmablasts" = "goldenrod4",
                     "Prolif Lymphocytes" = "#8fdbaf", "Megakaryocytes" = "darkolivegreen",
                     "HSPC" = "thistle1", "DC" = "orchid3",
                     "CD14 Monocytes" = "paleturquoise3", "CD16 Monocytes" = "#577676")
DimPlot(obj, reduction = "umap", group.by = "CellTypes", cols = celltype_colors, label = TRUE, repel = TRUE)
```

```{r umap-split, fig.width=12, fig.height=9}
DimPlot(obj, reduction = "umap", group.by = "CellTypes", cols = celltype_colors,
        split.by = "sample_id", ncol = 3) + NoLegend()
```

> **Exercise 5.2 (preview of Part 2).** Just by looking at the split UMAP,
> which cell types seem to change in COVID-19? Do any seem to differ between
> the female and the male samples? Write down your guess; in Part 2 we
> test it properly.

# 6. Save

We save a slim version of the object (counts, normalised data, UMAP and
metadata). Part 2 starts from this file, so students who could not finish
Part 1 can use the instructor's copy.

```{r save}
obj_small <- DietSeurat(obj, assays = "RNA", layers = c("counts", "data"), dimreducs = "umap")
if ("scale.data" %in% Layers(obj_small)) obj_small[["RNA"]]$scale.data <- NULL
obj_small@misc$gene_chr <- gene_chr
saveRDS(obj_small, file = file.path(params$out_dir, "D3_annotated_6samples.rds"))
```

```{r session-info}
sessionInfo()
```
